---
title: "Interest Rates"
author: "John C. Hull"
output:
  pdf_document:
    toc: true
---

# Chapter examples

## Example 5.1 - Forward price for an investment asset

$$
F_0 = S_0 e^{rT}\tag{5.1}
$$

Consider a 4-month forward contract to buy a zero-coupon bond that will mature 1
year from today. (This means that the bond will have 8 months to go when the
forward contract matures.) The current price of the bond is \$930. We assume
that the 4-month risk-free rate of interest (continuously compounded) is 6\% per
annum. Because zero-coupon bonds provide no income, we can use equation (5.1)
with $T = 4/12$, $r = 0.06$, and $S_0 = 930$. The forward price, $F_0$, is given
by

$$
F_0 = 930 e^{0.06 \times 4/12} = \$948.79
$$

---

## Example 5.2 - Known income

$$
F_0 = (S_0 - I)e^{rT}\tag{5.2}
$$

Consider a 10-month forward contract on a stock when the stock price is \$50. We
assume that the risk-free rate of interest (continuously compounded) is 8\% per
annum for all maturities and also that dividends of \$0.75 per share are
expected after 3 months, 6 months, and 9 months. The present value of the
dividends, $I$, is

$$
I = 0.75e^{-0.08 \times 3/12} + 0.75e^{-0.08 \times 6/12}
+ 0.75e^{-0.08 \times 9/12} = 2.162
$$

The variable $T$ is 10 months, so that the forward price, $F_0$, from
equation (5.2), is given by

$$
F_0 = (50 - 2.162)e^{0.08 \times 10/12} = \$51.14
$$

---

## Example 5.3 - Known yield

$$
R_c = m \ln{\left(1 + \frac{R_m}{m} \right)}\tag{4.3}
$$

$$
F_0 = S_0 e^{(r - q)T}\tag{5.3}
$$

Consider a 6-month forward contract on an asset that is expected to provide
income equal to 2\% of the asset price once during a 6-month period. The
risk-free rate of interest (with continuous compounding) is 10\% per annum. The
asset price is \$25. In this case, $S_0 = 25$, $r = 0.10$, and $T = 0.5$. The
yield is 4\% per annum with semiannual compounding. From equation (4.3), this is
3.96\% per annum with continuous compoudning. It follows that $q = 0.0396$, so
that from equation (5.3) the forward price, $F_0$, is given by

$$
F_0 = 25 e^{(0.10 - 0.0396) \times 0.5} = \$25.77
$$

---

## Example 5.4 - Valuing forward contracts

$$
F_0 = S_0 e^{rT}\tag{5.1}
$$

$$
f = (F_0 - K)e^{-rT}\tag{5.4}
$$

A long forward contract on a non-dividend-paying stock was entered into some
time ago. It currently has 6 months to maturity. The risk-free rate of itnerest
(with continuous compounding) is 10\% per annum, the stock price is \$25, and
the delivery price is \$24. In this case $S_0 = 25$, $r = 0.10$, $T = 0.5$, and
$K = 24$. From equation (5.1), the 6-month forward price, $F_0$, is given by

$$
F_0 = 25 e^{0.1 \times 0.5} = \$26.28
$$

From equation (5.4), the value of the forward contract is

$$
f = (26.28 - 24) e^{-0.1 \times 0.5} = \$2.17
$$

Equation (5.4) shows that we can value a long forward contract on an asset by
making the assumption that the price of the asset at the maturity of the forward
contract equals the forward price $F_0$.

---

## Example 5.5 - Futures prices of stock indices

$$
F_0 = S_0 e^{(r - q)T}\tag{5.8}
$$


Consider a 3-month futures contract on an index. Suppose that the stocks
underlying the index provide a dividend yield of 1\% per annum (continuously
compounded), that the current value of the index is 1,300, and that the
continuously compunded risk-free interest rate is 5\% per annum. In this case,
$r = 0.05$, $S_0 = 1,300$, $T = 0.25$, and $q = 0.01$. Hence, the futures price,
$F_0$, is given by equation (5.8)

$$
F_0 = 1,300 e^{(0.05 - 0.01)0.25} = \$1,313.07
$$

---

## Example 5.6 - Forward and futures contracts on currencies

$$
F_0 = S_0 e^{(r - r_f)T}\tag{5.9}
$$

Suppose that the 2-year interest rates in Australia and the United States are
3\% and 1\%, respectively, and the spot exchange rate is 0.7500 USD per AUD.
From equation (5.9), the 2-year forward exchange rate should be

$$
0.7500 e^{(.03 - .01)2} = 0.7206
$$

---

## Example 5.7 - Forward and futures contracts on currencies

$$
F_0 = S_0 e^{(r - r_f)T}\tag{5.9}
$$


In Table 5.4 on page 140, the September settlement price for the euro is
about 0.2\% higher than the September settlement price. This indicates that the
futures prices are increasing at about $4 \times 0.2 = 0.8\%$ per year with
maturity. From equation (5.9) this is a rough estimate of the amount by which
short-term euro interest rates were less than short-term U.S. itnerest rates in
May 2020.

---

## Example 5.8 - Futures on commodities

$$
F_0 = (S_0 + U)e^{rT}\tag{5.11}
$$

Consider a 1-year futures contract on an investment asset that provides no
income. It costs \$2 per unit to store the asset, with the payment being made at
the end of the year. Assume that the spot price is \$450 per unit and the
risk-free rate is 7\% per annum for all maturities. This corresponds to
$r = 0.07$, $S_0 = 450$, $T = 1$, and

$$
U = 2e^{-0.07 \times 1} = 1.865
$$

From equation (5.11), the theoretical futures price, $F_0$, is given by

$$
F_0 = (450 + 1.865)e^{0.07 \times 1} = \$484.63
$$

# Practice Questions

## 1. Forward price versus value of a forward contract

What is the difference between the forward price and the value of a forward
contract?

---

The forward price is the agreed-upon price at which the buyer and seller will
transact the underlying asset at a future date. It is determined at the
initiation of the contract and set so that the contract has zero value at that
moment. It depends on the spot price, interest rates, dividends (if any), and
time to maturity.

The value is how much the contract is worth to a party at a given time (can be
positive or negative). At initiation, thevalue is zero to both parties. Over
time, the value changes as market conditions (spot price, interest rates)
change. It is calculated as the difference between the current forward price and
the agreed forward price, discounted back to present value.

$$
\text{Value at time}\ t = [F_t(T) - K] \cdot e^{-r(T - t)}
$$

Where:

* $F_t(T)$: current forward price for delivery at time $T$
* $K$: contract price (original forward price)
* $r$: risk-free rate
* $T - t$: time remaining

### Summary

| Term | Meaning |
|:---|:---|
| Forward price | The price agreed at contract initiation for future delivery |
| Value of forward contract | The current worth of the contract to one party |

---

## 2. Forward price for an investment asset

Suppose that you enter into a 6-month forward contract on a non-dividend paying
stock when the stock price is \$30 and the risk-free interest rate (with
continuous compounding) is 5\% per annum. What is the forward price?

---

We can calculae the forward price using the standard formula for a non-dividend
paying stock with continuous compounding:

$$
F_0 = S_0 e^{rT}\tag{5.1}
$$

Where:

* $F_0$ is the forward price
* $S_0$ is the spot price (\$30)
* $r$ is the risk-free rate (0.05)
* $T$ is the time to maturity (6 months)

$$
F_0 = 30 e^{.05 \times .5} = 30.75945 \approx 30.76
$$

---

## 3. Futures prices of stock indices

A stock index currently stands at 350. The risk-free interest rate is 4\% per
annum (with continuous compounding) and the dividend yield on the index is 3\%
annum. What should the futures price for a 4-month contract be?

---

To calculate the futures price on a stock index that pays a continuous dividend
yield, we use the following formula under continuous compounding.

$$
F_0 = S_0 e^{r - q)T}\tag{5.8}
$$

Where:

* $F_0$ The future price
* $S_0$ the current index level (350)
* $r$ the risk-free rate (4\%)
* $q$ the dividend yield (3\%)
* $T = \frac{4}{12} = 0.333$ time to matuiry (4 months)

$$
F_0 = 350 e^{(0.04 - 0.03) \times .333} = 351.1674 \approx 351.17
$$

---

## 4. Futures on commodities

Explain carefully why the futures price of gold can be calculated from its spot
price and other observable variables whereas the futures price of copper cannot.

---

The difference between investment assets and consumption assets.

The ability to calculate a futures price from spot price and observable
variables like interest rates, storage costs, and convenience yield depends on
whether the commodity is easily and primarily held for investment or used for
consumption.

### Gold: Futures price is determinable

Gold is an investment asset that is often held as a store of value, and not
consumed. It can be stored, does not degrade, non-erishable and highly liquid.
Because of these traits, the futures price of gold can be derived using the
cost-of-carry model, which relies on observable variables:

$$
F_0 = S_0 e^{(r + u - y)T}
$$

Where:

* $S$ is the spot price
* $r$ is the risk-free interest rate
* $u$ storage cots (as a rate per unit)
* $y$ the convenience yield
* $T$ the time to maturity

Since all variables are observable or negligible, the futures price of gold is
deterministic and model-driven.

### Copper: Futres price is not fully determinable

Copper is a consumption commodity that is used in manufacturing, electronics, and
construction. It is costly to store and can degrade. Copper is also subject to
supply and demand uncertainty, which makes it valuable to hold for consumption.

The convenience yield for coppper can be significant and unobservable. It
reflects the non-monetary benefits of physically holding copper (e.g., avoiding
production delays). Since convenience yield is not directly observable or
constant, we can't precisely compute copper's futures price from spot price and
interest rates alone.

---

## 5. Known yield

Explain why a foreign currency can be treated as an asset providing a known
yield.

---

A foreign currency can be treated as an asset providing a known yield because
when you hold foreign currency, you have the opportunity to invest it in
risk-free instruments in the foreign country---such as Treasury bills or bank
deposits---and earn the foreign risk-free interest rate.

### Intuition

Suppose you hold 1,000 EUR. You can:

* Keep it as cash, or
* Deposit it in a eurozone bank or buy a euro-denominated bond.

Either way, you can earn the euro interest rate (say, 3\%). So, from the
perspective of a domestic investor, holding euros is similar to holding an asset
that pays a known yield of 3\%---just like a stock that pays a known dividend.

### Mathematical view

When pricing a forward contract on a foreign currency (say, EUR/USD), the
forward price is given by the cost-of-carry formula, treating the foreign
currency like a dividend-paying asset:

$$
F_0 = S_0 e^{(r - r_f)T}
$$

Where:

* $F_0$ is the forward exchange rate
* $S_0$ is the spot exchange rate
* $r$ domestic risk-free rate
* $r_f$ the foreign risk-free rate (the "known yield")
* $T$ the time to maturity

### Summary

A foreign currency earns its own interest rate in its home market, just like an
asset with a known yield. That's why, in financial modeling, we treat a foreign
currency as an investment asset providing a known yield equal to its risk-free
interest rate.

---

## 6. Futures pricing and expected values

Is the futures price of a stock index greater than or less than the expected
future value of the index? Explain your answer.

---

The futures price of a stock index is generally less than the expected future
value of the index. This is due to risk preferences and market pricing under the
risk-neutral measure rather than the real-world (or "true") probability
distribution.

### Here's the breakdown

#### 1. Futures price reflects the risk-neutral expectation, not the actual ("real-world") expected return.

* Future prices are determined by arbitrage---not by what investors expect the
  market to do, but by the cost of carry:

$$
F_0 = S_0 e^{(r - q)T}
$$

Where:

* $S$ the spot price
* $r$ the risk-free rate
* $q$ the dividend yield
* $T$ time to maturity
* This formula does not include any risk premium---it assumes investors are
  risk-neutral.

#### 2. Expected future value includes the equity risk premium

* In reality, the expected return on the stock index is usually higher than the
  risk-free rate (because investors demand compensation for risk).

* That means the expected future value of the index under the real-world measure
  includes this risk premium, and is higher than the futures price.

### Example

Assume:

* Spot index = 1,000
* Risk-free rate = 4\%
* Dividend yield = 2\%
* Time = 1 year

Then the futures price

$$
F_0 = 1,000 e^{(0.04 - 0.02)} \approx 1020.20
$$

But if the expected return on the index is 8\%, then the expected future value
of the index is:

$$
1,000 e^{0.08} \approx 1083.29
$$

So, the futures price is \$1020.20, which is less than the expected future
value \$1083.29.

### Conclusion

The futures price  of a stock index is usually less than the expected future
value of the index because it reflects a risk-neutral valuation, while the
expected value reflects investors' actual expectations including a risk premium.

---

## 7.  Valuing forward contracts

A 1-year long forward contract on a non-dividend-paying stock is entered into
when the stock price is \$40 and the risk-free rate of interest is 5\% per
annum with continuous compounding.

a. What are the forward price and the initial value of the forward contract?
b. Six months later, the price of the stock is \$45 and the risk-free
   interest rate is still 5\%. What are the forward price and the value of
   the forward contract?

---

### A. Forward price and initial value of the forward contract

From equation 5.1, the forward price ($F_0$) is:

$$
F_0 = 40 e^{.05 \times 1} = 42.05084 \approx 42.05
$$

Initialy, the forward contract has zero value to both parties.

### B. Six months later...

* Time remaining: $T = .5$
* New stock price: $S_0 = 45$
* Original forward price: 42.05

The new forward price:

$$
F_t = 45 e^{.05 \times .5} = 46.13918 \approx 46.14
$$

The new forward price (for a new contract) = \$46.14

The value of the contract:

$$
f = (46.14 - 42.05) e^{-.05 \times .5} = 3.989018 \approx 3.99
$$

---

## 8. Calculating the futures price of a stock index

The risk-free rate of interest is 7\% per annum with continuous compounding, and
the dividend yield on a stock index is 3.2\% per annum. The current value of the
index is 150. What is the 6-month futures price?

---

$$
F_0 = S_0 e^{(r-q)T} \tag{5.1}
$$

Where:

* $F_0$ is the futures price
* $S_0$ is the current index level ~ 150
* $r$ is the risk free rate ~ .07
* $q$ is the dividend yield ~ .032
* $T$ time ti maturity in years ~ .5

$$
F_0 = 150 \cdot e^{(.07 - .032)\cdot 0.5} \approx 152.88
$$

---

## 9. Variable dividend yield and stock index futures

Assume that the risk-free interest rate is 4\% per annum with continuous
compounding and that the dividend yield on a stock index varies throughout
the year. In February, May, August, and November, dividends are paid at a rate
of 5% per annum. In other months, dividends are paid at a rate of 2\% per annum.
Suppose that the value of the index on July 31 is 1,300. What is the futures
price for a contract deliverable in December 31 of the same year?

---

Time interval: July 31 - December 31

$$
\frac{5}{12}
$$

Adjusted Equation

$$
F_0 = S_0 e^{(r - \bar{q})T}
$$

The average annualized dividend yield:

$$
\bar{q} = \frac{ 2 \cdot 0.05 + 3 \cdot 0.02}{5} = 0.032
$$

$$
F_0 = 130 \cdot e^{(0.04 - 0.032)\cdot \frac{5}{12}} \approx 1304.34
$$

---

## 10. Index futures and arbitrage strategies

Suppose that the risk-free interest rate is 6\% per annum with continuous
compounding and that the dividend yield on a stock index is 4\% per annum. The
index is standing at 400, and the futures price for a contract deliverable in
four months is 405. What arbitrage opportunities does this create?

---

To determine whether there is an arbitrage opportunity, we need to:

1. Calculate the theoretical futures price based on spot price, interest rate,
   and dividend yield.

2. Compare it to the actual futures price.

3. If there's a discrepancy, evaluate whether casn-and-carry or reverse
   cash-and-carry arbitrage is possible.

### Step 1: Calculate the theoretical futures price

$$
F_{\text{theoretical}} = S_0 \cdot e^{(r - q)T}
$$

Where:

* $S_0 = 400$ is the spot index level.
* $r = 0.06$ is the risk-free rate.
* $q = 0.04$ is the dividend yield.
* $T = \frac{4}{12}$ is the time to maturity.

$$
F = 400 \cdot e^{(0.06 - 0.04)\cdot \frac{1}{3}} \approx 402.68
$$

### Step 2: Compare to actual futures price

* Actual futures price is 405
* Theoretical futures price is 402.68

The actual price is higher than the theoretical price and thus the arbitrage
opportunity is cash-and-carry arbitrage.

### Step 3: Execute arbitrage strategy

1. Borrow \$400 at 6\% per annum (continuously compounded) for four months.
2. Buy the index at 400.
3. Sell the futures contract at 405 (short the futures).
4. Over 4 months:

> * Receive dividends at 4\% per annum
$400 \cdot (1 - e^{-0.04 \cdot \frac{1}{3}}) \approx 5.28$
> * Repay loan $400 \cdot e^{0.06 \cdot \frac{1}{3}} \approx 408.08$

5. At futures maturity:

> * Deliver the index via the futures contract and receive 405.
> * Total inflow: 405 + 5.28 (dividends) = 410.28
> * Total outflow: repay loan = 408.08

Profit $\approx$ 410.28 - 408.08 = \$2.20 per index unit

### Final answer

There is an arbitrage opportunity because the actual futures price (405) is
above the theoretical price (402.68). You can execute a cash-and-carry arbitrage
by:

* Buying the index,
* Shorting the futures,
* Earning dividends and lcking in a risk-free profit of approximately \$2.20 per
  index unit.

---

## 11. Short-term interest rates from currency differentials

Estimate the difference between short-term interest rates in Japan and the
United States on May 21, 2020, from the information in Table 5.4 on page 140.

### Japanese Yen, USD per 100 yen, 12.5 million yen

| Date | Open | High | Low | Prior Settlement | Last trade | Change | Volume |
|:---|:---|:---|:---|:---|:---|:---|:---|:---|
|June 2020 | 0.93015 | 0.93035 | 0.92745 | 0.93070 | 0.92970 | -0.00100 | 61,018 |
| Sept. 2020 | 0.9340 | 0.93125 | 0.92895 | 0.93200 | 0.93125 | -0.00075 | 453 |

---

To estimate the difference between short-term interest rates in Japan and the
United States on May 21, 2020, we can use the prices of currency futures for
Japanese yen.

Each futures contract is on 12.5 million yen, quoted in USD per 100 yen. The
relevant futures prices:

* June 2020 contract: 0.92970
* September 2020 contract: 0.93125

Let's interpret this using the covered interest parity (CIP) condition, which
relates spot and forward exchange rates to interest rate differentials:

$$
F_0 = S_0 \cdot e^{(r_{\text{USD}} - r_{\text{JPY}})T}
$$

But since we don't have the spot exchange rate directly, we treat the June
contract as the spot proxy (it's very close to expiration), and use the
September contract as the forward.

### Step-by-step calculation

Let:

* $S_0 = 0.92970$ June 2020 spot
* $F_0 = 0.93125$ Sept 2020 forward price (quote)
* $T = 3/12 = 0.25$ years (June to September)

We use `R` to solve

```{r problem 11: calculate may futures}
S.0 <- 0.92970
F.0 <- 0.93125
T <- .25
f <- function(r.diff) S.0 * exp(r.diff * T) - F.0
uniroot(f, upper=1, lower=1e-4, tol=1e-8)$root
```

The difference between short-term interest rates in the U.S. and Japan on May
21, 2020, is approximately 0.667\% (U.S. higher)

This implies that short-term interest rates in the U.S. were about 0.667
percentage points higher than in Japan at the time.

---

## 12.

The 2-month interest rates in Switzerland and the United States are,
respectively, 1\% and 2\% per annum with continuous compounding. The spot price
of the Swiss franc is \$1.0500. The futures price for a contract deliverable in
2 months is \$1.0500. What arbitrage opportunities does this create?

---

Given:

* Swiss 2-month rate: $r_{\text{CHF}} = 1\%$
* U.S. 2-month rate: $r_{\text{USD}} = 2\%$
* Spot price of CHF: $S = 1.0500$ USD/CHF
* Futures price: $F = 1.0500$ USD/CHF
* Time to maturity: $T = 2/12 = 1/6 years$

All interest rates are continuously compounded.

### Step 1: What should the futures price be?

Using covered interest parity (CIP):

$$
F_0 = S_0 \cdot e^{(r_{\text{USD}} - r_{\text{CDF}})T} \approx 1.0518
$$

### Step 2: Compare to actual futures price

* Theoretical futures price: 1.0518
* Actual futures price: 1.0500

The actual price is too low relative to the theoretical price.

### Step 3: Arbitrage strategy -- Reverse cash-and-carry

To exploit undervalued futures:

1. Borrow CHF at 1\% per annum for 2 months.
2. Convert CHF to USD at spot rate $S = 1.0500$.
3. Invest USD at 2\% per annum for 2 months.
4. Enter into a long futures contract to buy CHF in 2 months at 1.0500.
5. At maturity:

  * Use proceeds from USD investment to buy CHF via the futures contract.
  * Repay CHF loan.

You've locked in a better forward rate than interest rate parity justifies, so
you earn riskless arbitrage profit.

### Final Answer

Yes, there is an arbitrage opportunity because the futures price (1.0500) is too
low relative to its theoretical value (1.05175).

The arbitrage involves:

* Borrowing Swiss francs
* Converting to U.S. dollars
* Investing in U.S. at higher interest
* Locking in a cheap futures contract to repurchase Swiss francs

This earns a risk-free profit due to the mispricing in the futures market.

---

## 13.

The spot price of silver is \$25 per ounce. The storage costs are \$0.24 per
ounce per year payable quarterly in advance. Assuming that interest rates are
5\% per annum for all maturities, calculate the futures price of silver for
delivery in 9 months.

---

To calculate the futures price of silver when there are storage costs, we adjust
the cost-of-carry formula accordingly.

### Given

* Spot price $S_0 = \$25$ per ounce
* Storage cost = \$0.24 per year, paid quarterly in advance
* Time to maturity: $T = \frac{9}{12} = 0.75$ years.
* Risk-free rate: $r = 5\% = 0.05$, with continuous compounding

### Step 1: Understand the storage costs

Quarterly payments = \$0.24/4 = \%0.06 per quarter.

Since they are paid in advance, the present value (PV) of each is discounted
from its payment time, not end of quarter.

Payments occur at:

* $t = 0$
* $t = 0.25$
* $t = 0.50$

(We ignore the 4th payment at $t = 0.75$ since the contract ends before the
next payment.)

Present value of storage costs:

$$
\begin{aligned}
PV(\text{storage}) &= 0.06 \cdot e^{-0.05 \cdot 0}
  + 0.06 \cdot e^{-0.05 \cdot 0.25} + 0.06 \cdot e^{-0.05 \cdot 0.05}\\
  &= 0.06 \cdot 2.9629\\
  &\approx 0.1777733
\end{aligned}
$$

### Step 2: Apply the cost-of-carry formula

$$
\begin{aligned}
F_0 &= (S_0 + PV(\text{storage}))\cdot e^{rT}\\
  &= (25 + 0.1778) \cdot e^{-.05 \cdot 0.75}\\
  &\approx 26.13
\end{aligned}
$$

The 9-month futures price of silver is approximatgely: \$26.13 per ounce.

---

## 14.

Suppose that $F_1$ and $F_2$ are two futures contracts on the same commodity
with times to maturity, $t_1$ and $t_2$, where $t_2 > t_1$. Prove that
$F_2 \le F_1 e^{r(t_2 - t_1)}$, where $r$ is the interest rate (assumed
constant) and there are no storage costs. For the purposes of this problem,
assume that a futures contract is the ssame as a forward contract.

---

We are asked to prove that for two futures (or forward) prices $F_1$ and $F_2$
with maturities $t_1$ and $t_2$ respectively, where $t_2 > t_1$, the following
inequality holds:

$$
F_2 \le F_1 e^{r(t_2 - t_1)}
$$

Assumptions:

* Both contracts are on the same commodity
* No storage costs
* Interest rate $r$ is constant (continuous compounding)
* Futures behave like forwards (i.e., delivery-style contracts with no interim
  cash flows)

### Intuition behind the inequality

This inequality asserts a no-arbitrage condition: If the futures price for a
longer-maturity contract were too high, arbitrageurs could lock in a riskless
profit by:

* Selling the longer-term contract (i.e., taking a short position),
* Buying the shorter-term contract and rolling it forwarrd,
* Earning the risk-free return in between.

### Step-by-Step proof

Suppose $F_2 > F_1 e^{r(t_2 - t_1)}$.

We'll show that this leads to an arbitrage opportunity, violating the
no-arbitrage principle.

### Arbitrage Strategy (Assume $F_2 > F_1 e^{r(t_2 - t_1)}$):

1. At time 0:

   * Enter a long position in the contract maturing at $t_1$ (agree to buy the
     commodity at $F_1$).
   * Enter a short position in the contract maturing at $t_2$ (agree to sell the
     commodity at $F_2$).
   * Borrow $F_1$ at the risk-free rate $r$ until $t_1$.

2. At time $t_1$:

   * Use borrowed funds $F_1 e^{r t_1}$ to buy the commodity via the long
     contract.
   * Hold the commodity from $t_1$ to $t_2$ (no storage cost).
   * No cost to holding.

3. At time $t_2$:

   * Deliver the commodity to fulfill the short futures contract, receiving
     $F_2$.

### Net cash flows

* Time 0: Net cash flow = 0 (only entering into contracts, borrowing $F_1$, no
  initial cost).
* Time $t_1$: Buy commodity for $F_1$, repay loan = $F_1 e^{r t_1}$
* Time $t_2$: Receive $F_2$

Thus, profit at $t_2$:

$$
\text{Profit} = F_2 - F_1 e^{r(t_2 - t_1)} \cdot e^{r t_1}
  = F_2 - F_1 e^{r t_2}
$$

But we assumed $F_2 > F_1 e^{r(t_2 - t_1)} \Rightarrow$ arbitrage profit > 0.

This contradicts the no-arbitrage condition, hence:

$$
F_2 \le F_1 e^{r(t_2 - t_1)}
$$

### Conclusion

Under the assumptions (no storage costs, constant interest rate, futures =
forwards), the no-arbitrage condition implies:

$$
F_2 \le F_1 e^{r(t_2 - t_1)}
$$

This inequality ensures there is no opportunity to profit from price
discrepancies between futures of different maturities.

---

## 15.

When a known future cash outflow in a foreign currency is hedged by a company
using a forward contract, there is no foreign exchange risk. When it is hedged
using futures contracts, the daily settlement process does leave the company
exposed to some risk. Explain the nature of this risk. In particular, consider
whether the company is better off using a futures contract or a forward contract
when:

a. The value of the foreign currency falls rapidly during the life of the
contract.

b. The value of the foreign currency rises rapidly during the life of the
contract.

c. The value of the foreign currency first rises and then falls back to its
initial value.

d. The value of the foreign currency first falls and then rises back to its
initial value.

Assume that the forward price equals the futures price.

---

When a company hedges a known future foreign currency cash outflow, such as a
payment it must make in a nother currency, it can use forward or futures
contracts. If forward and futures prices are equal, both lock in the same
exchange rate for the final settlement. However, futures contracts are marked to
amrket daily, meaning gains or losses are realized and settled each day, while
forward contracts settle only once at maturity.

Because of this, futures contracts introduce "margin risk" or "liquidity risk"
even when the overall hedge is economically sound. This risk arises from cash
flow timing differences between the hedge and the underlying exposure.

### Nature of the risk with futures

#### Forward contract:

No intermediate cash flows. The company is not required to post margin or deal
with mark-to-market volatility. The gain or loss is realized at contract
maturity.

#### Futures contracts:

Requires daily margining. Gains/losses from currency movements are realized
immediately, leading to potential cash flow mismatches, especially if the
foreign currency value reverses before contract maturity.

Now, let's analyze each scenario assuming the company needs to pay in the
foreign currency at contract maturity and uses a futures contract to hedge.

### The foreign currency falls rapidly during the life of the contract

* The futures position incurs losses early (because the company has a long
  futures position to hedge a future outflow).

* Dure to daily settlement, the company must post margin or realize cash losses
  early.

* These losses are not offset by lower payment in the foreign currency until
  later, at maturity.

* Foreard contract is better: No margin required; the loss is settled at the
  end.

### The foreign currency rises rapidly during the life of the contract

* The futures position shows gains early, and the company receives cash inflows
  from daily marking to market.

* These gains provide early liquidity before the actual outflow.

* These gains provide early liquidity before the actual outflow.

* Even though both forward and futures end with the same value, the futures
  hedge improves cash flow.

* Better to use a futures contract.

### The foreign currency first rises, then falls back to its initial value

* The company receives gains early but then returns them as the currency falls
  (daily losses).

* Net effect at maturity is zero, but cash inflows early followed by outflows
  later may cause liquidity strain.

* Forward contract avoids this churn.

* Better to use a forward contract.

### The foreign currency first falls, then rises back to its initial value

* The company experiences early losses (posts margin), then recovers gains
  later.

* The hedge is still economically correct at maturity, but the company may face
  temporary liquidity pressure.

* Better to use a forward contract.

### Concluison

| Scenario | Best hedge instrument | Reason |
|:---|:---|:---|
| Currency falls rapidly | Forward | Avoids early cash outflows from futures losses |
| Currency rises rapidly | Futures | Provides early gains and cash inflows |
| Currency rises then falls | Forward | Avoids unnecessary margin churn |
| Currency falls then rises | Forward | Avoids early liquidity drain |

Thus, for managing foreign exchange exposure with minimal liquidity risk,
forwards are generally preferred, except in cases where early gains from
favorable movements in futures might be beneficial.

---

## 16.

It is sometimes argued that a forward exchange rate is an unbiased predictor of
future exchange rates. Under what circumstances is this so?

---

The claim that a forward exchange rate is an unbiased predictor of future spot
exchange rates is grounded in expectations theory, and it holds only under
specific assumtpions.

### When is the forward rate an unbiased predictor?

The forward exchange rate is an unbiased predictor of the future spot rate if
the following conditions are met:

1. __Rational expectaions:__ Market participants form expectations about future
   spot rates rationally, using all available information.

2. __Risk neutrality:__ Investors are indifferent to risk; they do not require a
   risk premium for holding foreign exchange positions.

3. __Efficient markets:__ There are no arbitrage opportunities, and all known
   information is already reflected in prices.

4. __No transaction costs, taxes, or capital controls:__ These can distort
   arbitrage relationships and lead to deviations between forward rates and
   expected spot rates.

### What does "unbiased" mean?

The forward rate being unbiased means that:

$$
F_0 = E(S_t)
$$

Where:

* $E(S_t)$ is the expected spot rate at time $t$.
* $F_0$ is the forward rate agreed at initial time.

So, on average over time, the forward rate equals the future spot rate. It does
not mean it's a good predictor in every instance.

### When the forward rate is not unbiased

In the real world, the forward rate often deviates systematically from future
spot rates due to:

* __Risk premia:__ If investors are risk-averse, they may demand a premium for
  exposure to certain currencies. This leads to a bias in the forward rate.

* __Speculation and carry trades:__ Investors seeking higher interest rate
  differentials can distort forward pricing.

* __Behavioral biases and market inefficiencies__

### Empirical evidence

Empirical studies (like those on the "forward premium puzzle") have shown that
forward rates are often poor predictors of future spot rates, especially in the
short run and for high-interest-rate currencies.

---

## 17.

Show that the growth rate in an index futures price equals the excess return on
the portfolio underlying the index over the risk-free rate. Assume that the
risk-free interest rate and the dividend yield are constant.

---

To show that the growth rate in an index futures price equals the excess return
on the underlying portfolio over the risk-free rate, we assume:

* The risk-free rate $r$ is constant (continuously compounded),
* The dividend yield $q$ on the index is constant (also continuously
  compounded),
* The index futures price at time $t$ is $F(t)$.
* The spot price of the index at time $t$ is $S(t)$.

### Step 1: Futures pricing formula

The futures price of a stock index that pays a known continuous dividend yield
$q$ is given by:

$$
F(t) = S(t) e^{(r - q)(T -t)}
$$

Where:

* $T$ is the time to maturity of the futures contract,
* $(r - q)$ is the cost of carry (risk-free rate minus dividend yield).

This expression comes from no-arbitrage pricing: buying the index (and earning
dividends) while borrowing at the risk-free rate must match the futures payoff.

### Step 2: Compute the growth rate of the futures price

To find the instantaneous rate of change (growth rate) of the futures price,
take the derivative with respect to $t$:

$$
\frac{dF(t)}{dt} = \frac{d}{dt}\left[ S(t) e^{(r - q)(T - t)} \right]
$$

Applying the product rule:

$$
\frac{dF(t)}{dt} = \frac{dS(t)}{dt} e^{(r - q)(T - t)}
  + S(t) \cdot \frac{d}{dt} \left(e^{(r - q)(T - t)} \right)
$$

Now, compute the derivative of the exponential term:

$$
\frac{d}{dt} \left(e^{(r - q)(T - t)} \right) = - (r - q) e^{(r - q)(T - t)}
$$

So,

$$
\frac{dF(t)}{dt} = \frac{dS(t)}{dt} e^{(r - q)(T - t)}
  - (r - q) S(t) e^{(r - q)(T - t)}
$$

Factor out $e^{(r - q)(T - t)}$:

$$
\frac{dF(t)}{dt} = \left( \frac{dS(t)}{dt}
  - (r - q)S(t) \right) e^{(r - q)(T - t)}
$$

Now divide by $F(t) = S(t) e^{(r - q)(T - t)}$ to get the growth rate:

$$
\begin{aligned}
\frac{1}{F(t)} \cdot \frac{dF(t)}{dt} = \frac{1}{S(t)}
  \cdot \left(\frac{dS(t)}{dt} - (r - q)S(t) \right)\\
  \Rightarrow \frac{1}{F(t)} \cdot \frac{dF(t)}{dt}
  = \frac{1}{S(t)} \cdot \frac{dS(t)}{dt} - (r - q)
\end{aligned}
$$

This gives:

$$
\text{Growth rate of futures price} = \frac{1}{S(t)}
  \cdot \frac{dS(t)}{dt} - (r - q)
$$

But from the basic asset return formula:

$$
\frac{1}{S(t)} \cdot \frac{dS(t)}{dt} + q = \text{Total return on the index}
$$

So:

Growth rate of futures price = Return on the index - q - (r - q)
 = Return on the index - risk-free rate.

### Interpretation

This result shows that the rate at which the futures price increases reflects
how much the index outperforms the risk-free rate. If the index earns exactly
the risk-free rate, the futures price stays constant. If the index outperforms,
the futures price rises, if it underperforms, the futures price falls.

---

## 18.

Explain carefully what is meant by the expected price of a commodity on a
particular future date. Suppose that the futures price for crude oil declines
with the maturity of the contract at the rate of 2\% per year. Assume that
speculators tend to be short crude oil futures and hedgers tend to be long. What
does the Keynes and Hicks argument imply about the expected future price of oil?

---

### What is the expected price of a commodity on a future date?

The expected price of a commodity on a particular future date is the market's
consensus estimate--based on current information--of what the spot price of the
commodity will be at that future time. It reflect the average of all possible
future spot prices, weighted by their probabilities, as perceived today.

Mathematically, if $S_T$ is the spot price at future time $T$, then the expected
price is:

$$
E(S_T)
$$

This expectation is generally taken under the real-world (physical) probability
measure---not the risk-neutral measure used in pricing derivatives. It
incorporates market beliefs, supply and demand expectations, geopolitical risks,
and other real-world uncertainties.

### How do futures prices relate to expected prices?

In an ideal frictionless market with:

* no risk premiums,
* no storage costs or convenience yields,
* and no arbitrage opportunities,

the futures price $F(t, T)$ would equal the expected future spot price:

$$
F(t, T) = E(S_T)
$$

However, in reality, futures prices often differ from expected future spot
prices because of risk premiums. This is where the Keynes and Hicks theory comes
in.

### Keynes and Hicks' Theory: Normal backwardation

Keynes and Hicks (1930s) introduced the idea of normal backwardation, which
explains why futures prices may be lower than expected future spot prices.

* __Hedgers__ (e.g., oil producers) want to protect themselves from a fall in
  prices, so they take long positions in futures.

* __Speculators__ are willing to take the opposite short positions, but only if
  compensated for the risk.

* Therefore, futures prices are set below the expected future spot price to
  offer speculators a positive expected return.

This creates a risk premium in the futures market.

Mathematically:

$$
F(t, T) < E(S_T)
$$

### Application to crude oil futures

#### Assumptions:

* The futures price for crude oil declines with contract maturity at a rate of
  2\% per year.

* Speculators are short crude oil futures.

* Hedgers are long crude oil futures.

#### Interpretation:

If the futures curve is downward sloping (i.e., backwardation), and speculators
are net short, then--according to Keynes and Hicks--the futures price is below
the expected future spot price.

So, the market is offering speculators an expected profit (positive expected
return) for taking the risk of falling prices.

### Conclusion:

> The Keynes and Hicks argument implies that the expected future spot price of
> crude oil is higher than the futures price.

So, even though futures prices are declining with maturity (suggesting lower
future prices), the theory says this reflects a risk premium, not an unbiased
forecast. The actual expected spot price of oil in the future is above the
quoted futures price curve.

---

## 19.

The Value Line Index is designed to reflect changes in the value of a portfolio
of over 1,600 equally weighted stocks. Prior to March 9, 1988, the change in the
index from one day to the next was calculated as the geometric average of the
changes in the prices of the stocks underlying the index. In these
circumstances, does equation (5.8) correctly relate the futures price of the
index to its cash price? If not, does the equation overstate or understate the
futures price?

$$
F_0 = S_0 e^{(r - q)T}\tag{5.8}
$$

---

Equation 5.8 is the standard pricing formula for stock index futures. This
formula assumes that the index is a value-weighted portfolio of stocks and that
dividends are paid proportionally to the index weight (as with the S&P 500).

### Issue with geometric averaging (Value Line Index pre-1988)

Prior to March 9, 1988, the Value Line Index used the geometric average of the
price relatives of its ~1,600 equally weighted stocks:

$$
\text{Index}_t = \left( \prod_{i = 1}^n \frac{P_{i, t}}{P_{i,t-1}} \right)^{1/n}
$$

This geometric averaging method does not conform to the assumptions behind
Equation 5.8, which is derived under arbitrage arguments applicable to
value-weighted or replicable portfolios.

### Why equation 5.8 fails in this case

The geometric average of returns is generally lower than the arithmetic average
due to Jensen's inequality. Because geometric averaging reduces the effect of
high-return stocks, it results in a lower expected return compared to a
value-weighted index. The dividend yield $q$ is hard to define properly in this
context. Most importantly: you cannot replicate the index exactly with a
tradeable portfolio due to geometric weighting. Hence, no-arbitrage pricing
(used to derive equation 5.8) breaks down.

### Does equation 5.8 overstate or understate the futures price?

Because the geometric average tends to understate the growth rate of the index
compared to a value-weighted index, using equation 5.8 (which assumes
exponential growth at the rate $r - q$) will overstate the true futures price.

### Conclusion

> No, equation 5.8 does not correctly relate the futures price to the spot price
> of the Value Line Index before March 9, 1988. It overstates the futures price
> because it assumes a higher growth rate than what a geometric average of stock
> returns would actually produce.

This discrepency was one of the reasons why the Value Line Index changed its
calculation method in 1988 to an arithmetic average, making it more consistent
with traditional futures pricing models.

---

## 20.

A U.S. company is interested in using the futures contracts traded by the CME
Group to hedge its Australian dollar exposure. Define $r$ as the interest rate
(all maturities) on the U.S. dollar and $r_f$ as the interest rate (all
maturities) on the Australian dollar. Assume that $r$ and $r_f$ are constant and
that the company uses a contract expiring at time $T$ to hedge an exposure at
time $t$ ($T > t$).

a. Show that the optimal hedge ratio is $e^{(r_f - r)(T - t)}$, ignoring daily
   settlement.

b. Show that, when $t$ is 1 day, the optimal hedge ratio is almost exactly
   $S_0/F_0$, where $S_0$ is the current spot price of the currency and $F_0$ is
   the current futures price of the currency for the contract maturing at time
   $T$.

c. Show that the company can take account of the daily settlement of futures
   contracts for a hedge that lasts longer than 1 day by adjusting the hedge
   ratio so that it always equals the spot price of the currency divided by the
   futures price of the currency.

---

This question examines optimal hedging with foreign exchange futures,
particularly when the hedging horizon is shorter than the futures contract
maturity. We'll walk through each part.

Given:

* $r$ USD interest rate (continuously compounded)
* $r_f$ AUD interest rate (continuously compounded)
* $T$ expiration of the futures contract
* $t < T$ time of actual exposure
* $S_0$ current spot rate (USD per AUD)
* $F_0$ current futures price for delivery at $T$

### (a) Optimal hedge ratio $h^* = e^{(r_f - r)(T - t)}$

The goal is to find how much of the futures contract should be used to hedge a
foreign currency exposure maturing at $t$, using a futures contract that matures
at $T > t$.

#### Step-by-step:

1. The forward price $F_0$ for time $T$ is given by the cost-of-carry model:

$$
F_0 = S_0 e^{(r - r_f)T}
$$

2. Now consider the value of one unit of foreign currency (e.g., AUD 1) at time
   $t$, discounted from $T$. The present value at time $t$ of a short position
   in a futures contract maturing at $T$ is:

$$
V_t = F_0 e^{-r(T - t)} - S_t
$$

3. But we want to match this with the exposure in foreign currency, which grows
   at $r_f$. So we scale the hedge such that the foreign currency value at $t$
   is equivalent to the USD value delivered by the futures at $T$:

$$
h^* = \frac{e^{r_f(T - t)}}{e^{r(T - t)}} = e^{(r_f - r)(T - t)}
$$

#### Result

$$
h^* = e^{(r_f - r)(T - t)}
$$

### (b) When $t = 1$ day, show $h^* \approx \frac{S_0}{F_0}$

$$
h^* = e^{(r_f - r)(T - t)} \approx 1 + (r_f - r)(T - t)
$$

Also, recall:

$$
F_0 = S_0 e^{(r - r_f)T} \Rightarrow \frac{S_0}{F_0} = e^{(r_f - r)T}
$$

Now, if $T - t$ is very small (e.g., 1 day $\approx \frac{1}{365}$), then:

$$
h^* = e^{(r_f - r)(T - t)} \approx \frac{S_0}{F_0}
$$

Why? Because:

$$
\frac{S_0}{F_0} = e^{(r_f - r)T} = e^{(r_f - r)(T - t)} \cdot e^{(r_f - r)T}
  = h^* \cdot^{(r_f - r)t}
$$

If $t$ is small (1 day), then $e^{(r_f - r)t} \approx 1$, so:

$$
h^* \approx \frac{S_0}{F_0}
$$

### (c) Hedging with daily settlement: update hedge ratio to $\frac{S_t}{F_t}$

#### Problem:

Futures are marked to market daily. This means gains and losses are realized
before the final exposure date.

#### Insight:

To stay correctly hedged through time, you must rebalance the hedge so that the
ratio of spot to futures stays correct for the current state.

#### Solution:

Adjust the hedge ratio continuously as:

$$
h^*_t = \frac{S_t}{F_t}
$$

This ensures the hedge remains delta-neutral, even in the presence of daily cash
flows from futures.

---

## 21.

What is the cost of carry for:

a. a non-dividend-paying stock
b. a stock index
c. a commodity with storage costs
d. a foreign currency.

---

The cost of carry is the net cost or benefit of holding an asset until the
maturity of a forward or futures contract. It typically includes financing
costs, income (like dividends), and storage or convenience yields.

### a. Non-dividend-paying stock

* __Cost of carry:__ $c = r$

Since the stock pays no dividends, the only cost is financing the position
(e.g., borrowing money to buy the stock). Therefore, the cost of carry ($c$) is
the risk-free interest rate ($r$).

### b. Stock index

* __Cost of carry:__ $c = r - q$

The dividend yield ($q$) acts as a negative cost (i.e., a benefit) when holding
the index. Therefore, the cost of carry ($c$) is the risk-free interest rate
($r$) minus the dividend yield ($q$).

### c. Commodity with storage costs

* __Cost of carry:__ $c = r - u$

Commodities often incur storage costs, and may also involve insurance or
spoilage. If there's also a convenience yield ($y$), the formula becomes:
$c = r + u - y$

$y$ reflects the non-monetary benefit of physically holding the comodity (e.g.,
to ensure supply).

### d. Foreign currency

* __Cost of carry:__ $c = r - r_f$

Holding foreign currency yields interest at rate $r_f$, while the opportunity
cost of capital is $r$. So the net cost of carry is the difference.

---

## 22.

The spot exchange rate between the Swiss franc and U.S. dollar is 1.0404 (\$ per
franc). Interest rates int he United States and Switzerland are 0.25\% and 0\%
per annum, respectively, with continuous compounding. The 3-month forward
exchange rate was 1.0300 (\$ per franc). What arbitrage strategy was possible?
How does your answer change if the forward exchange rate is 1.0500 (\$ per
franc).

---

To determine the arbitrage strategy, we compare the theoretical forward rate
implied by interest rate parity with the actual forward rate quoted in the
market.

### Given:

* Spot rate, $S_0 = 1.0404$ (USD/CHF)
* U.S. interest rate, $r = 0.25\% = 0.0025$ per annum (continuous compounding)
* Swiss interest rate, $r_f = 0.00$
* Time to maturity, $T = 3$ months $= 0.25$ years
* Market forward rate $F_0 = 1.0300$ (USD/CHF)

### Step 1: Compute the theoretical forward rate

Using the interest rate parity formula (with continuous compounding):

$$
\begin{aligned}
F_0 &= S_0 \cdot e^{(r - r_f)T}\\
  &= 1.0404 e^{(0.0025 - 0.00)(0.25)}\\
  &\approx 1.04105
\end{aligned}
$$

### Case 1: Market forward rate = 1.0300 $\rightarrow$ Too low

You can profit by buying low and selling high in the forward market.

Arbitrage strategy:

1. Borrow CHF (e.g., 1 million CHF) at 0\% for 3 months.

2. Convert to USD at spot: $1,000,000 \times 1.0404 = 1,040,400$ USD.

3. Invest USD at 0.25\% for 3 months:
   $1,040,400 \cdot e^{.0025 \cdot .25} \approx 1,041,050$

4. Enter a forward contract to buy CHF in 3 months at 1.0300

5. In 3 months use your 1,041,050 USD to buy back CHF at 1.0300:
   $1,041,050/1.0300 \approx 1,010,729$ CHF

6. Repay 1 million CHF, and pocket the profit of 10,729 CHF.

This is covered interest arbitrage: No exchange rate risk and a guaranteed
profit.

### Case 2: Market forward rate = 1.0500 $\rightarrow$ Too high

Now you can sell high in the forward market and buy low in the spot.

Reverse arbitrage strategy:

1. Borrow USD (e.g., 1,040,400 USD) at 0.25\% for 3 months.

2. Convert to CHF at spot: $1,040,400/1.0404 = 1,000,000$ CHF

3. Invest CHF at 0\% for 3 months $1,000,000$ CHF

4. Enter a forward contract to sell CHF for USD at 1.0500.

5. In 3 months, deliver 1,000,000 CHF for:
   $1,000,000 \times 1.0500 = 1,050,000$ USD

6. Repay USD loan: $1,040,400 \cdot e^{.000625} \approx 1,041,050$

7. Profit $1,050,000 - 1,041,050 = 8,950$ USD

---

## 23. Futures price of an index

An index is 1,200. The three-month risk-free rate is 3\% per annum and the
dividend yield over the next three months is 1.2\% per annum. The six-month
risk-free rate is 3.5\% per annum and the dividend yield over the next six
months is 1\% per annum. Estimate the futures price of the index for three-month
and six-month contracts. All interest rates and dividend yields are continuously
compounded.

---

$$
F_0 = S_0 \cdot e^{(r - q)T}
$$

### Where:

* $F_0$: futures price
* $S_0$: current index level = 1,200
* $r$: risk-free interest rate (continuously compounded)
* $q$: dividend yield (continuously compounded)
* $T$: time to maturity (in years)

### 1. Three-month futures price

Given:

* $r = 0.03$
* $q = 0.012$
* $T = \frac{3}{12} = 0.25$

$$
F_0^{(3m)} = 1200 \cdot e^{(0.03 - 0.012)\cdot 0.25} \approx 1,205.41
$$

### 2. Six-month futures price

Given:

* $r = 0.035$
* $q = 0.01$
* $T = \frac{6}{12} = 0.5$

$$
F_0^{(6m)} = 1200 \cdot e^{(0.035 - 0.01)\cdot 0.5} \approx 1,215.09
$$

---

## 24. 

Suppose the current USD/euro exchange rate is 1.2000 dollar per euro. The
six-month forward exchange rate is 1.1950. The six-month USD interest rate is
1\% per annum continuously compouned. Estimate the six-month euro interest rate.

---

$$
F_0 = S_0 e^{(r - r_f)T}
$$

$$
\begin{aligned}
1.1950 = 1.2000 \cdot e^{(.01 - r_f)T}\\
e^{(.01 - r_f)\cdot .5} = \frac{1.1950}{1.2000}\\
(r_f - .01)\cdot .5 = \ln{\left( \frac{1.1950}{1.2000} \right)}\\
r_f = .01 - 2 \cdot \ln{\left( \frac{1.1950}{1.2000} \right)}\\
r = 0.01835
\end{aligned}
$$

This can be solved via `R`

```{r problem 24}
F.0 <- 1.1950
S.0 <- 1.2000
r <- .01
T <- .5
f <- function(r.f) S.0 * exp((r - r.f)*T) - F.0
(uniroot(f, upper=1, lower=1e-4, tol=1e-8)$root)
```

---

## 25.

The spot price of oil is \$50 per barrel and the cost of storing a barrel of oil
for one year is \$3, payable at the end of the year. The risk-free interest rate
is 5\% per annum continuously compounded. What is an upper bound for the
one-year futures price of oil?

---

Given:

* $S_0$: Spot price = 50
* $r$: risk-free rate = .05
* $U$ = present value of storage costs
* $T$ = Time to maturity = 1 year

$$
F_0 > (S_0 + U) \cdot e^{rT}
$$

The present value of the storage costs ($U$):

$$
\begin{aligned}
U &= 3 \cdot e^{-rT}\\
  &= 3 \cdot e^{-.05}\\
  &= 2.8537
\end{aligned}
$$

The upper bound of the futures price is:

$$
\begin{aligned}
F_{\text{upper bound}} &= (50 + 2.8537) \cdot e^{.05}\\
  &= 55.56
\end{aligned}
$$

---

## 26.

A company that is uncertain about the exact date when it will pay or receive a
foreign currency may try to negotiate with its bank a forward contract that
specifies a period during which delivery can be made. The company wants to
reserve the right to choose the exact delivery date to fit in with its own cash
flows. Put yourself in the position of the bank. How would you price the product
that the company wants?

---

If you're a bank pricing a window forward contract (a forward contract with a
delivery window instead of a fixed delivery date), you must account for the
flexibility it gives the client. This flexibility adds value for the client and
risk for the bank, so the price (forward rate) must be adjusted accordingly.

---

### Key considerations for the bank

#### Client optionality:

The clien can choose the most favorable delivery date within the window. This is
similar to giving them a free option---they'll deliver (or take delivery) when
the forward rate is most advantageous.

#### Hedging challenge:

The bank cannot perfectly hedge the exposure because the exact date is
uncertain. This introduces basis risk.

#### Volatility and time window:

The longer and more volatile the window, the more optionality it gives the
client---and the more costly it is to hedge.

#### Midpoint vs. Worst point pricing:

The bank may price the forward based on: The midpoint of the forward rates over
the window if risk-sharing). Or, more conservatively, the least favorable (for
the bank) forward rate in the window to compensate for the optionality.

### Pricing approaches

#### Worst-case forward rate:

Price the forward at the least favorable rate (for the bank) over the delivery
window:

$$
\text{Forward rate} = \text{min}\{F(t_1), F(t_2), \ldots, F(t_n)\}
$$

or

$$
\text{Forward rate} = \text{max}\{F(t_1), F(t_2), \ldots, F(t_n)\}
$$

depending on whether the client is buying or selling the currency.

#### Average forward plus optionality premium:

Compute the average forward rate over the window and add a premium to reflect
the cost of optionality (possible estimated using options pricing models like
Black-Scholes).

### Bottom line

To price the product, the bank would typically either:

* Quote a conservative forward rate that protects against the most adverse date
  the client could choose, or

* Charge an additional premium or spread to compensate for the optionality the
  client receives.

This turns the product into a structured forward contract, which blends elements
of forwards and options.

---

## 27.

A company enters into a forward contract with a bank to sell a foreign currency
for $K_1$ at time $T_1$. The exchange rate at time $T_1$ proves to be $S_1$
($> K_1$). The company asks the bank if it can roll the contract forward until
time $T_2$ ($> T_1$) rather than settle at time $T_1$. The bank agrees to a new
delivery price, $K_2$. Explain how $K_2$ should be calculated.

---

When a company wants to roll a forward contract from time $T_1$ to a later time
$T_2$, the bank essentially replaces the original contract with a new forward
contract. The new forward price $K_2$ must reflect:

* The current spot or forward rate at time $T_1$,

* The time value of money over the new interval $[T_1, T_2]$,

* The fact that the original contract had an in-the-money or out-of-the-money
  value at time $T_1$.

### Given:

* Original forward contract to sell foreign currency at price $K_1$ at time
  $T_1$,

* At time $T_1$, the spot exchange rate is $S_1$ and $S_1 > K_1$,

* New forward contract desired for settlement at $T_2 > T_1$ with new delivery
  price $K_2$.

### Step-by-step: How the bank should calculate $K_2$

#### 1. Value the original contract at $T_1$:

Since the company had a contract to sell at $K_1$, and the spot is $S_1$, the
value of the contract at $T_1$ is:

$$
V_1 = K_1 - S_1\ \ \text{(from the company's perspective, this is a loss)}
$$

But this is from the bank's perspective (bank will lose money if it honors the
contract at the old rate). To be fair, the value should be based on the forward
price at time $T_1$, not the spot price.

So more accurately:

$$
V_1 = K_1 - F(T_1, T_1) = K_1 - S_1
$$

if we assume the forward price equals the spot price at maturity (no arbitrage).

#### 2. Calculate new forward price $K_2$:

The new delivery price $K_2$ must:

* Reflect the current forward price from $T_1$ to $T_2$, say $F(T_1, T_2)$,

* Adjust for the gain or loss on the original contract (i.e., the bank should
  recoup any loss it would have realized at $T_1$).

Thus, the fair adjustment leads to:

$$
K_2 = F(T_1, T_2) + \frac{V_1}{e^{r(T_2 - T_1)}}
$$

Here, $r$ is the domestic risk-free rate and the term
$\frac{V_1}{e^{r(T_2 - T_1)}}$ discounts the bank's loss (or gain over the new
forward period.

### Final formula

$$
K_2 = F(T_1, T_2) + \frac{K_1 - S_1}{e^{r(T_2 - T_1)}}
$$

This means the new forward price equals the market forward rate over the
extended horizon plus the present value of the difference between the original
contract price and the spot rate at the roll time.

### Interpretation

* If the original forward was out of the money for the bank (i.e., $K_1 < S_1$),
  the bank adds a premium to the new forward price $K_2$. 

* If the original forward was in the money, $K_2$ might be below the new market
  forward price.

This ensures no arbitrage and fair compensation for both parties in the
roll-forward arrangement.

---
