---
title: Options
output: pdf_document
---

# Options trading

## Selling a call option (Short call)

* You are selling someone the right to buy a stock from you at a specific price
  (strike price) before a certain date (expiration date).

* You collect a premium upfront.

* Outcomes:

  1. If out of the money you collect the premium.

  2. If in the money, you collect premium and sell the stock at the strike
     price. If the stock sells higher than the stock price you miss out on the
     stock's gain.

Example:

  * Strike price: \$10
  * Premium: \$1

```{r sell call option, echo = F}
strike <- 10
premium <- 1
S_T <- seq(0, 20, by = 1)
profit <- premium - pmax(0, S_T - strike)
plot(S_T, profit, type = "l", col = "blue", lwd = 1,
     ylab = "Profit", xlab = "Stick price at expiration",
     main = "Profit from selling a call option")
abline(h = 0, lty = 2)
text(1, .5,  "premium")
text(8, .5,  "strike price")
text(13.5, - .5,  "break even")
```

## Buying a call option (Long call)

* You are buying the right to buy a stock from someone at a specific price
  (strike price) before a certain date (expiration date).

* You pay a premium upfront for the option.

* Outcomes:

  1. If in the money you have the right to buy the stock at the discounted
     strike price but you pay a premium. Your payout is based on the stock price
     minus the premium you paid.

  2. If out of money the option expires worthless. You lose just the premium.

Example:

* Strike price: \$10
* Premium: \$1

```{r buy call option, echo = F}
strike <- 10
premium <- 1
S_T <- seq(0, 20, by = 1)

profit <- pmax(0, S_T - strike) - premium

plot(S_T, profit, type = "l", col = "blue",
     ylab = "Profit", xlab = "Stock price at expiration",
     main = "Buying a call option")
abline(h = 0, lty = 2)
text(1,-.5, "premium")
text(8, -.5, "strike price")
text(14, .5, "break even")
```

## Selling a put option (Short put)

* You are selling someone the right to sell a stock to you for a specific price
  (strike price) before a certain date (expiration date).

* You collect a premium upfront.

* Outcomes:

  1. If out of the money you collect the premium.

  2. If in the money, you collect the premium and buy the stock at the strike
     price. If the stock sells lower than the stock price you are paying more
     for the stock than what its worth.

Example:

* Strike price: \$10
* Premium: \$1

```{r sell put option, echo = F}
strike <- 10
premium <- 1
S_T <- seq(0, 20, by = 1)

profit <- premium - pmax(0, strike - S_T)

plot(S_T, profit, type = "l", col = "blue",
     ylab = "Profit", xlab = "Stock price at expiration",
     main = "Selling a put option")
abline(h = 0, lty = 2)
text(19, .5, "premium")
text(12, .5, "strike price")
text(6, -.5, "break even")
```

## Buying a put option

* You are buying the right to sell someone a stock at a specific price (strike
  price) before a certain date (expiration date).

* You pay a premium upfront.

* Outcomes:

  1. If out of the money you pay the premium and the option expires worthless.

  2. If in the money you can sell the stock at the strike price. Your profit is
     the strike price minus the current price minus the premium.

Example:

* Strike price: \$10

* Premium: \$1

```{r buy put option, echo = F}
strike <- 10
premium <- 1
S_T <- seq(0, 20, by = 1)

profit <- pmax(0, strike - S_T) - premium

plot(S_T, profit, type = "l", col = "blue",
     ylab = "Profit", xlab = "Stock price at expiration",
     main = "Buying a put option")
abline(h = 0, lty = 2)
text(19, -.5, "premium")
text(12, -.5, "strike price")
text(6, .5, "break even")
```

