---
title: "Introduction"
author: "John C. Hull"
#date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
---

```{r setup, include=FALSE}
require(latex2exp)
require(dplyr)
require(kableExtra)
options(scipen = 999)
knitr::opts_chunk$set(echo = TRUE)
yen_symbol <- "&#xA5;"
pretty <- function(x) format(x, bigmark=",")
prettyd <- function(x) format(x, "%B %d, %Y")
text <- new.env()
```

# Chapter Highlights

## 1.1 Exchange-traded markets

A _derivative exchange_ is a market where individuals and companies trade
standardized contracts that have been defined by the exchange.

The CBOT and CME have merged to form the CME group.

* Chicago Board of Trade (CBOT) 1848

* Chicago Mercantile Exchange (CME) 1919

* New York Mercantile Exchange (NYMEX)

* Kansas City Board of Trade (KCBT)

* The Chicago Board Options Exchange (CBOE) 1973

Once two traders have agreed to trade a product offered by an exchange, it is
handled by the _exchange clearing house_. The clearing house manages the risk
between the two traders.

---

## 1.2 Over-the-counter markets

Banks and other large financial institutions such as fund managers, insurance
agencies, and large corporations are the main participants in OTC derivatives
markets.

Once an OTC trade has been agreed, the two parties can either present it to a
central coutnerparty (CCP) or clear the trade bilaterally.

The over-the-counter market is significantly bigger than the exchange traded
market for derivatives.

However, the OTC market has not grown much since 2007 because of the popularity
of a procedure called _compression_. Compression is where two or more
counterparties restructure transactions with each other with the result that the
underlying principal is reduced.

---

## 1.3 Forward contracts

A _forward contract_ is an agreement to buy or sell an asset at a certain future
time for a certain price. A _spot contract_ is an agreement to buy or sell an
asset almost immediately.

Forward contracts are traded in over-the-counter markets.

The _long position_ agrees to buy the underlying asset at a certain time and
price.

The _short position_ agrees to sell the underlying asset at a certain time and
price.

### Payoffs from forward contracts

#### Long position:

$S_T - K$

```{r payoffs from forward contracts; long, echo=F, fig.width=6, fig.height=4}
S.T <- 0:100
payoff <- S.T - 50
plot(S.T, payoff, type='l', axes=F, xlab='', ylab='')
arrows(0, 0, 100, 0, length=.05)
arrows(0, -50, 0, 53, length=.05)
text(50, 0, 'K', pos=1, cex=.8)
text(100, 0, expression(S[T]), pos=1, cex=.8)
text(0, 50, 'Payoff', pos=4, cex=.6)
```

#### Short position:

$K - S_T$

```{r payoffs from forward contracts; short, echo=F, fig.width=6, fig.height=4}
S.T <- 0:100
payoff <- 50 - S.T
plot(S.T, payoff, type='l', axes=F, xlab='', ylab='')
arrows(0, 0, 100, 0, length=.05)
arrows(0, -53, 0, 53, length=.05)
text(50, 0, 'K', pos=1, cex=.8)
text(100, 0, expression(S[T]), pos=1, cex=.8)
text(0, 50, 'Payoff', pos=4, cex=.6)
```

Where:

* $S_T$ is the spot price of the asset at maturity of the contract.
* $K$ is the delivery price.

### Forward prices and spot prices

Consider a stock that pays no dividend and is worth \$60. You can borrow or lend
money for 1 year at 5\%. What should the 1-year forward price of the stock be?

The answer is \$60 grossed up at 5\% for 1 year, or \$63. If the forward price
is more than this say \$67, you could borrow \$60, buy one share of the stock,
and sell it forward for \$67. After paying off the loan, you would net a profit
of \$4 in 1 year. If the forward price is less than \$63, say \$58, an investor
owning the stock as part of a portfolio would sell the stock for \$60 and enter
into a forward contract to buy it back for \$58 in 1 year. The proceeds of
investment would be invested at 5\% to earn \$3. The investor would end up \$5
better off than if the sotck were in the portfolio for the year.

---

## 1.4 Futures contracts

Similar to forward contracts, a _futures contract_ is an agreement between two
parties to buy or sell an asset at a certain time in the future for a certain
price. However, futures contracts are traded on an exchange.

To make trading possible, the exchange speicifies certain standardized features
of the contract.

Two large exchanges on which futures contracts are traded are the Chicago Board
of Trade (CBOT) and the Chicago Mercantile Exchange (CME), which have now merged
to form the CME Group.

On these exchanges a wide variety of commodities and financial assets form the
underlying assets in the various contracts.

The commodities include pork bellies, live cattle, sugar, wool, lumber, copper,
aluminum, gold, and tin.

The financial assets include stock indices, currencies, and Treasury bonds.

---

## 1.5 Options

Options are traded both on exchanges and in the over-the-counter market.

Two types of options:

* _Call option_ gives the holder the right to buy the underlying asset by a
  certain date for a certain price.

* _Put option_ gives the holder the right to sell the underlying asset by a
  certain date for a certain price.

The price in the contract is known as the _exercise price_ or _strike price_.

The date in the contract is known as the _expiration date_ or _maturity_.

_American options_ can be exercised at any time up to the expiration date.

_European options_ can be exercised only on the expiration date itself.

The largest exchange in the world for trading stock options is the Chicago Board
Options Exchange (CBOE).

There are four types of participants in options markets:

1. Buyers of calls
2. Sellers of calls
3. Buyers of puts
4. Sellers of puts

Buyers are referred to as having long positions and sellers are referred to
having short positions.

Selling an option is also known as writing the option.

---

## Types of traders

There are three broad categories of traders: hedger, speculators, and
arbitrageurs. Hedgers use derivatives to reduce the risk that they face from
potential future movements in a market variable. Speculators use them to bet on
the future direction of a market variable. Arbitrageurs take offsetting
positions in two or more instruments to lock in a profit.

---

## Hedgers

Suppose that it is `r prettyd(text$current.date <- as.Date("2020-05-21"))`, and
ImportCo, a company based in the United States, knows that it will have to pay
10 million pounds on `r prettyd(text$future.date <- as.Date("2020-08-21"))`, for
goods it has purchased from a British supplier. The GBP/USD exchange rate quotes
made by a financial institution are shown in Table 1.1. Import Co could hedge
its foreign exchange risk by buying pounds (GBP) from the financial institution
in the 3-month forward market at 1.2225. This would have the effect of fixing
the price to be paid to the British exporter at \$12,225,000.

---

## Speculators

---

## Arbitrageurs

---

## Dangers
