---
title: Hedging Strategies Using Futures
output:
    pdf_document:
        toc: true
---

# Chapter examples

## Example 3.1

It is March 1. A U.S. company expects to receive 50 million Japanese yen at the
end of July. Yen futures contracts on the CME Group have delivery months of
March, June, September, and December. One contract is for the delivery of 12.5
million yen. The company therefore shorts four September yen futures contracts
on March 1. When the yen are received at the end of July, the company closes out
its position. We suppose that the futures price on March 1 in cents per yen is
1.0800 and that the spot and futures prices when the contract is closed out are
1.0200 and 1.0250 respectively.

The gain on the futures contract is 1.0800 - 1.0250 = 0.0550 cents per yen. The
basis is 1.0200 - 1.0250 = -0.0050 cents per yen when the contract is closed
out. The effective price obtained in cents per yen is the final spot price plus
the gian on the futures:

$$
1.0200 + 0.0550 = 1.0750
$$

This can also be written as the initial futures price plus the final basis:

$$
1.0800 + )-0.0050) = 1.0750
$$

The total amount received by the company for the 50 million yen is
$50 \times 0.01075$ million dollars, or \$537,500.

---

## Example 3.2

It is June 8 and a company knows that it will need to purchase 20,000 barrels of
crude oil at some time in October or November. Oil futures contracts are
currently traded for delivery every month by the CME Group and the contract size
is 1,000 barrels. The company therefore decides to use the December contract for
hedging and takes a long position in 20 December contracts. The futures price on
June 8 is \$48.00 per barrel. The company finds that it is ready to purchase the
crude oil on November 10. It therefore closes out its futures contract on that
date. The spot price and futures price on November 10 are \$50.00 per barrel and
\$49.10 per barrel.

The gain on the futures contract is 49.10 - 48.00 = \$1.10 per barrel. The bases
when the cotnract is closed out is 50.00 - 49.10 = \$0.90 per barrel. The
effective price paid (in dollars per barrel) is the final spot price less the
gain on the futures, or

$$
50.00 - 1.10 = 48.90
$$

This can also be calculated as the initial futures price plus the final basis,

$$
48.00 + 0.90 = 48.90
$$

The total price paid is $48.90 \times 20,000 = \$978,000$.

---

## Example 3.3

An airline expects to purchase 2 million gallons of jet fuel in 1 month and
decides to use heating oil futures for hedging. We suppose that Table 3.2 gives,
for 15 successive months, data on the change, $\Delta S$, is the jet fuel price
per gallion and the corresponding change, $\Delta F$, in the futures price for
the contract on heating oil that would be used for hedgin price changes during
the month. In this case, the usual formulas for calculating standard deviations
and correlations give $\sigma_F = 0.0313$, $\sigma_S = 0.0263$, and 
$\rho = 0.928$.

```{r example 3.3: table 3.2, echo=F}
month <- c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14, 15)
delta.futures <- c(0.021, 0.035, -0.046, 0.001, 0.044, -0.029, -0.026, -0.029,
                   0.048, -0.006, -0.036, -0.011, 0.019, -0.027, 0.029)
delta.spot <- c(0.029, 0.020, -0.044, 0.008, 0.026, -0.019, -0.010, -0.007,
                0.043, 0.011, -0.036, -0.018, 0.009, -0.032, 0.023)
knitr::kable(data.frame(Month = month, DFutures = delta.futures,
                        DSpot = delta.spot),
             caption = "Data to calculate minimum variance hedge ratio")
```

$$
h^* = \rho \frac{\sigma_S}{\sigma_F}\tag{3.1}
$$

$$
N^* = \frac{h^* Q_A}{Q_F}\tag{3.2}
$$

```{r example 3.3: calculating standard deviation and correlation, echo = F}
cat("Standard deviation for spot:", sd(delta.spot), "\n",
    "Standard deviation for futures:", sd(delta.futures), "\n",
    "Correlation:", cor(delta.futures, delta.spot), "\n")
```
From equation (3.1), the minimum variance hedge ratio, $h^*$, is therefore

$$
0.928 \times \frac{0.0263}{0.0313} = 0.78
$$

Each heating oil contract traded by the CME Group is on 42,000 gallons of
heating oil. From equation (3.2), the optimal number of contracts is

$$
\frac{0.78 \times 2,000,000}{42,000} \approx 37
$$

which is 37 when rounded to the nearest whole number.

# Practice Questions

## 1. Perfect hedge

Explain what is meant by a _perfect hedge_. Does a perfect hedge always lead to
a better outcome than an imperfect hedge? Explain your answer.

---

A perfect hedge is one that completely eliminates the risk. Perfect hedges are
rate. For the most part, a study of hedging using futures contracts is a study
of the ways in which hedges can be constructed so that they perform as close to
perfectly as possible.

A perfect hedge does not always lead to better outcomes, because it may
eliminate upside potential and be costlier or less flexible than an imperfect
hedge. The choice between perfect and imperfect hedges depends on risk
tolerance, objectives, and cost-benefit trade-offs.

---

## 2. Minimum variance hedge

Under what circumstances does a minimum variance hedge portfolio lead to no
hedging at all?

---

A minimum variance hedge portfolio aims to reduce the variance (volatility) of
the value of a position by finding the hedge ratio that minimizes risk.
However, it can lead to no hedging at all under certain conditions.

The conditions under which a minimum variance hedge portfolio leads to no
hedging occur when:

a. No correlation between the spot (underlying) asset and the hedging instrument
(e.g., futures contract).

> * If the correlation coefficient ($\rho$) between the changes in the spot
    price and the hedge instrument is zero, then hedging provides no risk
    reduction.

> * In that case, the minimum variance hedge ration $(h^*) = 0$.

$$
h^* = \rho \cdot \frac{\sigma_S}{\sigma_F} = 0
$$

> Where:

> * $h^*$ is the optimal hedge ratio
> * $\rho$ is the correlation between spot and futures price changes
> * $\sigma_S, \sigma_F$ are the standard deviations of spot and futures price
    changes.

> If $\rho = 0$, then $h^* = 0$ no hedge.

b. Hedging instrument is ineffective: If the futures or other derivative does
not track the risk source well, the hedge might actually introduce more
variablility.

c. Costs outweigh benefits: Even if there's some correlation, high transaction
costs or margin requirements may make it optimal to not hege if the variance
reduction is minimal.

A minimum variance hedge portfolio leads to no hedging when the spot and hedge
instrument returns are uncorrelated ($\rho = 0$). In such cases, hedging
wouldn't reduce risk and might even worsen outcomes.

---

## 3. Optimal hedge ratio

Suppose that the standard deviation of quarterly changes in the prices of a
commodity is \$0.65, the standard deviation of quarterly changes in a futures
price on the commodity is \$0.81, and the coefficient of correlation between the
two changes is 0.8. What is the optimal hedge ratio for a 3-month contract? What
does it mean?

---

### Given:

* $\rho = 0.8$
* $\sigma_S = \$0.65$
* $\sigma_F = \$0.81$

$$
\begin{aligned}
h^* &= \rho \cdot \frac{\sigma_S}{\sigma_F}\\
  &= 0.8 \cdot \frac{0.65}{0.81}\\
  &\approx 0.642
\end{aligned}
$$

### Interpretation:
  
* The optima hedge ratio is approximately 0.642, or 64.2\%.
* This means that for every \$1 of exposure in the spot market, you should
  hedge \$0.642 using futures to minimize risk.
* In practical terms, if you're managing a position in the physical
  commodity, you'd short (or long) a futures position sized at 64.2\% of the
  spot exposure.

---

## 4. Hedging an equity portfolio and changing the portfolio beta

A company has a \$20 million portfolio with a beta of 1.2. It would like to use
futures contracts on a stock index to hedge its risk. The index futures price is
currently standing at 1080, and each contract is for delivery of \$250 times the
index. What is the hedge that minimizes risk? What should the company do if it
wants to reduce the beta of the portfolio to 0.6?

---

### Given:

* Portfolio value: \$20,000,000
* Beta: 1.2
* Index futures price: \$1080
* \$250 times index

### Hedge to minimize risk (i.e., reduce beta to 0)

$$
\begin{aligned}
N &= \frac{\beta \cdot V}{F \cdot Q}\\\\
N &= \frac{1.2 \cdot 20,000,000}{1080 \cdot 250}\\
  &= 88.89
\end{aligned}
$$

The company should short 89 index futures contracts to fully hedge (beta = 0)
and minimize risk.

### Hedge to reduce beta to 0.6

$$
\begin{aligned}
N &= \frac{(\beta - \beta^*) \cdot V}{F \cdot Q}\\\\
N &= \frac{(1.2 - 0.6) \cdot 20,000,000}{1080 \cdot 250}\\
  &= 44.44
\end{aligned}
$$

To reduce the portfolio beta to 0.6, the company should short 44 or 45 index
futures contracts.

---

## 5. Minimizing basis risk

In the corn futures contract traded on an exchange, the following delivery
months are available: March, May, July, September, and December. Which of the
available contracts should be used for hedging when the expiration of the hedge
is in (a) June, (b) July, and (c) January.

---

Choose the futures contract with an expiration date that is just after the
expiration of the hedge. This minimizes basis risk---the risk that the spot and
futures prices won't move perfectly in sync.

### Hedge expires in June

* Choose the July contract

### Hedge expires in July

* Choose the July contract

### Expires in January

* Choose the March contract

---

## 6. Perfect hedge

Does a perfect hedge always succeed in locking in the current spot price of an
asset for a future transaction? Explain your answer.

---

A perfect hedge does not always lock in the current spot price, even though it
eliminates price risk.

### What a perfect hedge does do

A perfect hedge ensures that the net outcome is known in advance, regardless of
what happens to prices. It locks in a known future net price, but that price may
be different from today's spot price.

### Why it might not lock in the current spot price

#### Futures prices $\ne$ spot prices today

A futures contract locks in the futures price, which may include interest rates,
storage costs, or expectations. If the futures price is higher or lower htan the
current spot price, your locked-in price reflects that.

#### Basis risk (for imperfect timing)

If the hedge doesn't mature exactly when the transaction occurs, the difference
between spot and futures (the basis) might change unpredictably. Even with a
hedge in place, you might not realize today's spot price due to basis movement.

### Summary

A perfect hedge eliminates price risk but does not necessarily lock in the
current spot price. It locks in the futures price, which may differ from the
spot price due to market conditions, interest rates, or time value.

---

## 7. The basis in hedging

Explain why a short hedger's position improves when the basis strengthens
unexpectedly and worsens when the basis weakens unexpectedly.

---

### What is the basis?

The _basis_ in a hedging situation is:

$$
\text{Basis} = \text{Spot price} - \text{Futures price}
$$

If the asset to be hedged and the asset underlying the futures contract are the
same, the basis should be zero at the expiration of the futures contract.

#### A short hedger (like a producer planning to sell an asset later)

Sells a futures contract now to lock in a selling price. Hopes to protect
against falling spot prices.

#### When the basis strengthens unexpectedly

That means the spot price falls less than the futures price (or even rises). So,
although the spot price might drop, the futures price drops more, creating a
bigger gain on the short futures position.

As a result, spot revenue is better than expected and futures position profits
more than expected. The hedger is better off overall.

#### When the basis weakens unexpectedly

The spot price drops more than the futures price. The futures position doesn't
offset as much of the spot price loss as expected.

As a result, spot revenue is worse than expected and futures gain is smaller.
The hedger is worse off overall.

---

## 8. Hedging foreign exchange risk

Imagine you are the treasurer of a Japanese company exporting electronic
equipment to the United States. Discuss how you would design a foreign exchange
hedging strategy and the arguments you would use to sell the strategy to your
fellow executives.

---

The objective is to protect the company's revenue (in JPY) from potential losses
due to a weakening USD (i.e., when USD/JPY declines).

### Step-by-step: Designing the FX hedging strategy

#### 1. Identify the exposure

* Sales are in USD, but costs are in JPY.
* You're exposed to USD depreciation against the JPY.

#### 2. Choose hedging instruments

| Instrument | How it works | Notes |
|:---|:---|:---|
| Forward contracts | Lock in exchange rate now for future settlement | Simple, flexible terms |
| FX options | Right (not obligation) to exchange at a specific rate | Useful if you want upside potential |
| Natural hedge | Match costs and revenues in the same currency | Harder for exporters |
| Currency swaps | Exchange cash flows over time | More complex, used for long-term hedging |

#### 3. Hedge the amount and timing

* Match hedge size to expected USD revenue over time.
* Consider using rolling hedges if cash flows are continuous.

#### 4. Monitor and adjust

* Regularly re-evaluate hedges as forecasts and market conditions change.

### Selling the strategy to executives

#### 1. Risk reduction

"We're protecting predictable cash flows. FX volatility can wipe out 5--10\% of
profits overnight."

#### 2. Earnings stability

"This gives us more predictable quarterly results---and more investor
confidence."

#### 3. Cost vs. benefit

"hedging is relatively low-cost insurance against large potential losses."

#### 4. Strategic flexibility

"We'll start with simple hedging (like forwards), and can adapt as we grow."

### Summary of your pitch

"Our exposure to USD/JPY puts millions in yen revenue at risk. A forward
contract hedge is simple, effective, and helps us lock in known outcomes, which
is vital in a volatile FX environment. With this strategy, we protect margins,
reduce earnings volatility, and demonstrate fiscal responsibility to
stakeholders."

---

## 9. Using the hedge ratio to determine optimal risk exposure

Suppose that in Example 3.2 above the company decides to use a hedge ratio of
0.8. How does the decision affect the way in which the hedge is implemented and
the result?

---

If the company decides to use a hedge ratio of 0.8 then it is choosing to hedge
only 80\% of its exposure using futures.

### What is the hedge ratio?

A hedge ratio of 0.8 means the company is only hedging:

$$
0.8 \times 20,000\ \text{barrels} = 16,000\ \text{barrels}
$$

Since each futures contract covers 1,000 barrels, the company now takes a long
position in 16 contracts instead of 20.

### How is the hedge implemented differently?

* __Original hedge:__ 20 futures contracts (100\% hedge)
* __ With hedge ratio of 0.8:__ 16 futures contracts (80\% hedge)
* Everything else (dates, prices, close-out timing) remains the same.

### Recalculate the result

#### Gain on futures:

$$
\begin{aligned}
49.10 - 48.00 = 1.10\ \text{per barrel}\\
\text{Total gain} = 1.10 \times 16,000 = \$17,600
\end{aligned}
$$

#### Spot market purchase:

On Nov 10, the company buys 20,000 barrels at \$50.00 = \$1,000,000

#### Net effective cost:

$$
\begin{aligned}
1,000,000 - 17,600 = 982,400\\
\text{Effective price per barrel} = \frac{982,400}{20,000} = \$49.12
\end{aligned}
$$

### Compared to full hedge

| Strategy | Effective price/barrel | Total cost
|:---|:---|:---|
| Full hedge | \$48.90 | \$978,000 |
| 80\% hedge | \$49.12 | \$982,400 |

### Interpretation

The partial hedge reduced risk, but not as much as a full hedge. The effective
purchase price is higher than with a full hedge because less gain was realized
from the futures contracts. However, the company might have chosen a hedge ratio
of 0.8 to leave room for potentially favorable price movements (e.g., if oil
prices fell).

### Summary

Using a hedge ratio of 0.8 means the company is only hedging 80\% of its
exposure and is taking on some risk, but possibly preserves more upside. Pays an
effective price of \$49.12 per barrel instead of \$48.90.

---

## 10. Minimum variance hedge ratio and the perfect hedge

"If the minimum variance hedge ratio is calculated as 1.0, the hedge must be
perfect." Is this statement true? Explain your answer.

---

The statement is not necessarily true.

### What is the minimum variance hedge ratio (MVHR)?

It's calculated as

$$
h^* = \frac{\text{Cov}(\Delta S, \Delta F)}{\text{Var}(\Delta F)}
$$

Where:

* $\Delta S$: change in spot price
* $\Delta F$: change in futures price

It tells you the number of futures contracts per unit of exposure that minimizes
variance (i.e., risk) of your position.

### What does $h^* = 1.0$ mean?

It means that the best risk-reducing strategy is to hedge the entire position
one-for-one using futures---i.e., for every dollar of exposure, take one
dollar's worth of futures.

However a perfect hedge would eliminate all risk---meaning that:

$$
\Delta S = \Delta F
$$

In practive the spot and futures prices don't move perfectly together. This
means there is basis risk (the spot-futures relationship changes unpredictably).
So even with $h^* = 1$, the hedge still has residual risk.

---

## 11. Basis risk and minimum variance hedge ratio

"If there is no basis risk, the minimum variance hedge ratio is always 1.0." Is
this statement true? Explain this statement.

---

This statement is true. If there is no basis risk then the futures price and the
spot price move perfectly together, meaning

$$
\Delta S = \Delta F
$$

There correlation is 1 and the variance of their difference is 0.

$$
\begin{aligned}
h^* = \frac{\text{Cov}(\Delta S, \Delta F)}{\text{Var}(\Delta F)}\\
\text{If}\ \Delta S = \Delta F\text{, then:}\\
\text{Cov}(\Delta S, \Delta F) = \text{Var}(\Delta F)
\end{aligned}
$$

Therefore, $h^* = 1$.

---

## 12. Future prices less than the spot price for long hedges

"When the futures price of an asset is less than the spot price, long hedges are
likely to be particularly attractive." Explain this statement.

---

A long hedge is when you expect to buy an asset in the future and want to
protect against price increases. You do this by buying futures contracts
now---locking in a future purchase price.

When the futures price < spot price, this situation is known as backwardation.
This means you can lock in buying the asset in the future for less than it costs
today. For a buyer, this looks like a discount---which makes the hedge very
appealing.

If the market stays in backwardation you lock in favorable pricing now. The
futures you boght may gain in value as they converge upward toward the spot
price at expiration. So, long hedgers benefit both from protection and potential
gain.

### One caveat

You're still not guaranteed to come out ahead---if spot prices fall, you might
have been better off waiting. But if your goal is risk reduction, and you can
lock in a cheaper-than-market price, a long hedge is very attractive.

---

## 13. Futures hedging strategies using MVHR

The standard deviation of monthly changes in the spot price of live cattle is
(in cents per pound) 1.2. The standard deviation of monthly changes in the
futures price of live cattle for the closest contract is 1.4. The correlation
between the futures price changes and the spot price changes is 0.7. It is now
October 15. A beef producer is committed to purchasing 200,000 pounds of live
cattle on November 15. The producer wants to use the December live cattle
futures contracts to hedge its risk. Each contract is for the delivery of 40,000
pounds of cattle. What strategy should the beef producer follow?

---

To hedge the price risk of purchasing 200,000 pounds of live cattle in one month
(November 15), the beef producer can use December live cattle futures.

### Step 1. User the minimum variance hedge ratio

$$
h^* = \rho \times \frac{\sigma_S}{\sigma_F}
$$

Where:

* $h^*$: hedge ratio (optimal number of futures per unit of spot).
* $\rho$: 0.7 (correlation between spot and futures price changes)
* $\sigma_S$: 1.2 (standard deviation of spot price changes)
* $\sigma_F$: 1.4 (standard deviation of futures price changes)

$$
h^* = 0.7 \times \frac{1.2}{1.4} = 0.7 \times 0.8571 \approx 0.6
$$

### Step 2. Determine the number of contracts to hedge

* Each futures contract = 40,000 pounds
* Total quantity to hedge = 200,000 pounds

#### Number of contracts

$$
\begin{aligned}
\text{Contracts} &= h^* \times \frac{\text{quantity to hedge}}
  {\text{contract size}}\\
  &= 0.6 \times \frac{200,000}{40,000}\\
  &= 0.6 \times 5\\
  &= 3\ \text{contracts}
\end{aligned}
$$

### Recommended strategy

The beef producer should take a long position in 3 December futures contracts on
October 15. This position should be closed on or around November 15 when the
cattle are purchased.

### Why a long hedge?

The producer is going to buy cattle, so they are exposed to the risk that prices
might rise. A long futures position offsets that risk by locking in a price (or
at least mitigating variance due to price changes).

---

## 14. Price risk versus production risk

A corn farmer argues "I do not use futures contracts for hedging. My real risk
is not the price of corn. It is that my whole crop gets wiped out by the
weather." Discuss this viewpoint. Should the farmer estimate his or her expected
production of corn and hedge to try to lock in a price for expected production?

---

This farmer is essentially saying that: If a natural disaster or drought
destroys the crop, there's nothing to sell, so the price becomes irrelevant.
Hedging the price would lock in a sale, but if there's no corn, the farmer would
still owe on the hedge (especially in a futures contract).

If the farmer hedges using futures contracts and cannot deliver due to crop
loss: They must close the futures position, possibly at a loss.

### Should the farmer hedge?

Yes, but strategically

#### Estimate expected production

Even if the total crop might be lost, total loss is rare. The farmer can: Use
historical yield data and weather forecasts to make an expected production
estimate. Hedge a portion of that expected yield---say 50-70\%, depending on
confidence in the crop.

#### Use flexible hedging instruments

Options (like buying a put) can lock in a floor price without obligating
delivery. If the crop is wiped out, the option simply expires---no obligation to
deliver, and only the premium is lost.

### Key ideas to sell to the farmer

1. Price risk is still real---even if production is uncertain.
2. A partial hedge can protect income from normal yields, without
   overcommitting.
3. Options can provide downside protection without delivery risk.
4. Crop insurance can be paired with hedging for broader risk management.


### Conclusion

The farmer is right to be cautious, but completely ignoring price risk is risky
too. A balanced approach---hedging part of the expected crop and/or using
options---offers the best of both worlds.

---

## 15. Heging market risk of a stock portfolio using beta hedging

On July 1, an investor holds 50,000 shares of a certain stock. The market price
is \$30 per share. The investor is interested in hedging against movements in
the market over the next month and decides to use an index futures contract. The
index futures price is currently 1,500 and one contract is for delivery of \$50
times the index. The beta of the stock is 1.3. What strategy should the investor
follow? Under what circumstances will it be profitable?

---

### Given

* 50,000 shares
* Stock price: \$30
* Value of stock $50,000\ \text{shares} \times \$30 = \$1,500,000$
* Use index futures contracts for hedging price movements
* Index futures price: 1,500
* \$50 times the index
* Value of futures contract: $1,500 \times \$50 = \$75,000$
* $\beta$: 1.3

### Step 1: Determine the hedge ratio (number of contracts)

$$
\begin{aligned}
\text{hedge ratio} &= \frac{\beta \times \text{value of portfolio}}
  {\text{value of one futures contract}}\\
  &= \frac{1.3 \times 1,500,000}{75,000}\\
  &= 26\ \text{contracts}
\end{aligned}
$$

Sell 26 index futures contracts to hedge. This is a short hedge, because the
investor is protecting against downside market risk.

### Under what circumstances will it be profitable?

This hedge is not designed to be profitable per se, but rather reduce risk.
However, the hedge is effective when:

* The market falls, causing the index futures to decline in value.
* Losses in the stock portfolio are offset by gains on the short futures
  position.

The hedge will underperform when:

* The market rises and the stock follows.
* The investor misses out on some (or all) upside gains due to losses on the
  futures.

---

## 16. Over-hedging

Suppose that in Table 3.5, the company decides to use a hedge ratio of
1.5. How does this affect the way the hedge is implemented and the result?

```{r problem 16: table 3.5, echo=F}
apr.2021 <- c(48.20, NA, NA, 49.00)
sept.2021 <- c(47.40, 47.00, NA, NA)
feb.2022 <- c(NA, 46.50, 46.30, NA)
june.2022 <- c(NA, NA, 45.90, 46.00)

df <- data.frame(apr.2021, sept.2021, feb.2022, june.2022)
rownames(df) <- c("Oct. 2021 futures price", "Mar. 2022 futures price",
                  "July 2022 futures price", "Spot price")

knitr::kable(df, caption = "Data for rolling oil hedge forward.")
```

---

By using a hedge ratio of 1.5 (instead of 1.0) means the company is
over-hedging: it is taking a futures position larger than the expected spot
transaction.

### Step 1: What does a 1.5 hedge ratio mean?

The company will short 150 contracts (instead of 100), because:

$$
\text{Hedge ratio} = \frac{\text{futures position}}
  {\text{spot exposure}} = 1.5
$$

* Each contract = 1,000 barrels
* Exposure = 100,000 barrels $\rightarrow$ 150 contracts = 150,000 barrels

So the hedge volume is larger than the actual oil to be sold.

### Step 2: Futures trading timeline

The company:

* Shorts 150 October 2021 contracts in April 2021 at \$48.20
* Closes them in September 2021 at \$47.40
* Shorts 150 March 2022 contracts in September 2021 at \$47.00
* Closes them in February 2022 at \$46.50
* Shorts 150 July 2022 contracts in February 2022 at \$46.30
* Closes them in June 2022 at \$45.90

### Step 3: Calculate gains from futures

Each round-trip futures trade gives a gain or loss per barrel:

1. April-Sept (Oct contract):

$$
48.20 - 47.40 = +\$0.80\ \text{per barrel}
$$

2. Sept-Feb (Mar contract):

$$
47.00 - 46.50 = +\$0.50
$$

3. Feb-June (July contract):

$$
46.30 - 45.90 = +\$0.40
$$

Total gain per barrel = 0.80 + 0.50 + 0.40 = \$1.70

but now the company is short 150,000 barrels in futures contracts, while it only
sells 100,000 barrels in the spot market

Total gain from futures:

$$
1.70 \times 150,000 = \$255,000
$$

## Step 4: Spot sale price

In June 2022, the spot price is \$46.00. So revenue from selling 100,000 barrels
is:

$$
100,000 \times 46.00 = \$4,600,000
$$

## Step 5: Total effective revenue

Add the gain from futures to the spot revenue:

$$
4,600,000 + 255,000 = \$4,855,000
$$

So, effective price per barrel = $\frac{4,855,000}{100,000} = \$48.55$

### Summary: How does a 1.5 hedge ratio change things?

#### Implementation changes

More contracts are traded at each stage: 150 vs. 100. The company is exposed to
basis risk on extra 50,000 barrels (over-hedged volume)

#### Result changes

Higher total gain from futures. Effective selling price increases, but at the
cost of over-hedging risk.

If prices moved against the futures position (i.e., prices rose), the over-hedge
would have amplified the loss.

---

## 17. Purpose of hedging using futures

An airline executive has argued: "There is no point in our using oil futures.
There is just as much chance that the price of oil in the future will be less
than the futures price as there is that it will be greater than this price."
Discuss the executive's viewpoint.

---

The problem with the executives position is that it treats the use of futures
contracts as a form of speculation, assuming the purpose is to "guess" future
prices correctly. But that is not what hedging is about.

### What hedging actually does

Hedging is not about making a profit---it's about reducing uncertainty. An
airline's primary concern is fuel cost volatility, since oil is a major input.
Using oil futures allows the airline to lock in a price for fuel in the future,
providing predictability in budgeting and financial planning.

If prices rise, the airline gains on the futures contract, offsetting higher
fuel costs. If prices fall, the airline loses on the futures contract, but pays
less for actual fuel--again, the net effect is stable.

The real value is reducing risk.

### Why hedging makes sense for airlines

Fuel is one of the largest and most volatile costs for airlines. Airlines
operate on thin profit margins and make forward-looing pricing decisions (e.g.
ticket sales months ahead). Locking in prices through futures helps stabilize
operating costs.

---

## 18. Calculating the forward price of a hedge

Suppose that the 1-year gold lease rate is 1.5\% and the 1-year risk-free rate
is 5.0\%. Both rates are compounded annually. Use the discussion in Business
Snapshot 3.1 to calculate the maximum 1-year gold forward price Goldman Sachs
should quote to the gold-mining company when the spot price is \$1,200.

---

The solution given in the business snapshot states that Goldman Sachs should
borrow the gold from a central bank, sell it immediately, and invest the
proceeds at the risk-free rate. At the end of the year, you buy the gold from
the gold mining company and use it to repay the central bank. The fixed forward
price you set for the gold reflects the risk-free rate you can earn and the
lease rate you pay the central bank for borrowing the gold.

### Given:

* Gold lease rate: .015
* Risk free-rate: .05
* Comounding: Annual
* Time: 1 year
* Spot price: \$1,200

### Strategy

1. Borrow from the central bank at .015 lease rate
2. Sell the gold immediately at the spot price \$1,200
3. Invest the proceeds at the risk-free rate (5.0\%)
4. At maturity, buy the gold from the mining company at the forward price $F$
   and use the gold to repay the loan

$$
\text{Future value} = 1,200 \times (1 + 0.05) = 1,260
$$

To break even, the cost of buying 1.015 oz of gold must be no more than \$1,260.

$$
\begin{aligned}
F \times 1.015 &= 1,260\\
F &= \frac{1,260}{1.015}\\
  &= 1,241.57
\end{aligned}
$$

The maximum 1-year forward price Goldman Sachs should quote is \$1,251,57 per
ounce.

---

## 19. Capital Asset Pricing Model (CAPM)

The expected return on the S&P 500 is 12\% and the risk-free rate is 5\%. What
is the expected return on an investment with a beta of (a) 0.2, (b) 0.5, and (c)
1.4?

---

We use the Capital Asset Pricing Model.

$$
\text{Expected return} = r_m + \beta (r_m - r_f)\\
$$

### For $\beta = 0.2$

$$
R = 5 + .2 (12 - 5) = 6.4\%
$$

### For $\beta = 0.5$

$$
R = 5 + .5 (12 - 5) = 8.5\%
$$

### For $\beta = 1.4$

$$
R = 5 + 1.4 (12 - 5) = 14.8
$$

---

## 20. Hedging futures contract strategies

It is now June. A company knows that it will sell 5,000 barrels of crude oil in
September. It uses the October CME Group futures contract to hedge the price it
will receive. Each contract is on 1,000 barrels of "light sweet crude." What
position should it take? What price risks is it still exposed to after taking
the position?

---

The company should short futures contracts and buy 5 contracts (1,000 barrels
per contract, 5,000 barrels total). This locks in a future selling price close
to the futures price today, reducing the risk of a price drop before September.

Even with the hedge, the company is still exposed to basis risk---this is the
risk that the spot price and the futures price don't move in perfect alignment.

### Reasons for basis risk:

#### Mismatch in timing

The hedge uses an October contract, but the oil is being sold in September,
before the contract matures. This means the company must close out the hedge in
September, and if the basis (spot price - futures price) behaves unexpectedly,
the hedge may not fully offset losses or gains in the spot price.

#### Quality/location mismatch

The hedge assumes the company's oil is similar to "light sweet crude" as
specified in the futures contract. If there are differences in grade or delivery
location, the hedge may not be perfect.

---

## 21. Basis risk and hedging position (short/long)

Sixty futures contracts are used to hedge an exposure to the price of silver.
Each futures contract is on 5,000 ounces of silver. At the time the hedge is
closed out, the basis is \$0.20 per ounce. What is the effect of the basis on
the hedger's financial position if (a) the trader is hedging the purchase of
silver and (b) the trader is hedging the sale of silver?

---

### A. Purchase of silver

When you are buying silver, you're worried the price may rise, so you go long
futures.

If the basis is positive (spot price > futures price), then you pay more in the
spot market than the price you lock in via futures and your futures gain offsets
part of the increased spot cost, but not all of it.

This would be a loss for the hedger due to the basis being positive at close
out.

$$
\text{Loss} = 6 \times 5,000 \times \$0.2 = \$60,000
$$

### B. Sale of silver

When you are selling silver, you're worried the price may fall, so you go short
futures.

If the basis is positive (spot price > futures price), then you sell in the spot
market at a higher price than the futures. You lose some value on your futures
position, but gain more in the spot sale.

This would be a gain for the hedger.

$$
\text{Gain} = 6 \times 5,000 \times \$0.2 = \$60,000
$$

---

## 22. Minimum variance hedge ratio and the number of contracts

A trader owns 55,000 units of a particular asset and decides to hedge the value
of her position with futures contracts on another related asset. Each futures
contract is on 5,000 units. The spot price of the asset that is owned is \$28
and the standard deviation of the change in this price over the life of the
hedge is estimated to be \$0.43. The futures price of the related asset is \$27
and the standard deviation of the change in this over the life of the hedge is
\$0.40. The coefficient of correlation between the spot price change and futures
price change is 0.95.

(a) What is the minimum variance hedge ratio?
(b) Should the hedger take a long or short position?
(c) What is the optimal number of futures contracts when adjustments for daily
settlement are not considered?
(d) How can the daily settlement of futures contracts be taken into account?

---

### Minimum variance hedge ratio

$$
\begin{aligned}
h^* &= \rho \frac{\sigma_S}{\sigma_F}\\
  &= 0.95 \frac{.43}{.40}\\
  &= 1.02125
\end{aligned}
$$

### Should the hedger take a long or short position?

The trader owns the asset and would like to hedge the value of the position. The
trader should take a short position. This way, if the value of the position
drops, the trader will benefit from the short position.

### What is the optimal number of futures contracts (no daily settlements)

The size of the position is

$$
Q_A = 55,000 \times 5,000 = 275,000
$$

$$
\begin{aligned}
N^* &= \frac{h^* Q_A}{Q_F}\\
  &= \frac{1.02125 \cdot 55,000}{5,000}
  &= 11.23 \approx 11\ \text{contracts}
\end{aligned}
$$

### How can the daily settlement of futures contracts be taken into account?

Daily settlemen (also called marking to market) affects the hedge because gains
or losses from the futures position are realized daily and can be withdrawn or
require margin top-up. This introduces the following considerations:

* __Cash flow timing risk:__ Since futures are settled daily but the spot aset
  may not be sold or realized until later, there may be a mismatch in cash
  flows.
* __Reinvestment risk:__ Gains from futures may need to be reinvested at
  uncertain rates.
* __Margin calls:__ Adverse futures movements may require posting additional
  collateral before the hedge matures.

To account for this:

* Firms may track and manage margin balances and consider them in cash
  budgeting.
* Some may use futures with lower volatility or options if cash flow stability
  is critial.

---

## 23. Cross hedging

A company wishes to hedge its exposure to a new fuel whose price changes have a
0.6 correlation with gasoline futures price changes. The company will lose \$1
million for each 1 cent increase in the price per gallon of the new fuel over
the next three months. The new fuel's price changes have a standard deviation
that is 50\% greater than price changes in gasoline futures prices. If gasoline
futures are used to hedge the exposure, what should the hedge ratio be? What is
the company's exposure measured in gallons of the new fuel? What position,
measured in gallons, should the company take in gasoline futures? How many
gasoline futures contracts should be traded? Each contract is on 42,000 gallons.

---

### Given:

* Asset correlation: 0.6
* \$1,000,000 per 1 cent change in price of new fuel
* SD of new fuel 50% greater than gasoline SD.

### Hedge ratio

$$
h^* = .6 \frac{1.5}{1} = .9
$$

### Company's exposure

The company loses \$1,000,000 for each 1 cent increase per gallon in the price
of the new fuel. A 1 cent increase per gallon isn \$0.01 per gallon. 

$$
Q = \frac{1,000,000}{0.01} = 100,000,000\ \text{gallons}
$$

The company's exposure is 100 million gallons of new fuel.

### Determine the effective position using gasoline futures

The hedge ratio of 0.9 means that for every gallon of new fuel exposure, the
hedging instrument (gasoline futures) should cover 0.9 gallons of exposure. So,
the hedge should cover:

$$
\text{Hedged exposure} = 0.9 \times 100,000,000 = 90,000,000\ \text{gallons}
$$

### Calculate the number of gasoline futures contracts needed

Each gasoline futures contract is on 42,000 gallons. Therefore, the number of
contracts required is:

$$
\begin{aligned}
\text{Number of contracts} &= \frac{\text{hedged exposure}}
  {\text{contract size}}\\
  &= \frac{90,000,000}{42,000}\\
  &\approx 2,142.86
\end{aligned}
$$

### Summary of results

* Hedge ratio: $h^* = 0.9$
* Total exposure: 100,000,000 of new fuel
* Futures position in gallons: The hedge should cover 90,000,000 gallons
* Number of gasoline futures contracts 2,143

### Interpretation

#### Hedge ratio of 0.9

Because the new fuel's price volatility is 50\% higher than that of gasoline
futures and the corelation between the two is 0.6, the optimal hedge is for 90\%
of the exposure.

#### Overall position

The company should take a short position in approximately 2,143 gasoline futures
contracts (each covering 42,000 gallons) to hedge its price risk on the new fuel
exposure.

This position will offset most, though not all, of the loss incurred from
adverse price movements in the new fuel, given the imperfect correlation and
different volatilities.

---

## 24. Capital Asset Pricing Model and portfolio performance

A portfolio manager has maintained an actively managed portfolio with a beta of
0.2. During the last year, the risk-free rate was 5\% and equities performed
very badly providing a return of -30\%. The portfolio manager produced a return
of -10\% and claims that in the circumstances it was a good performance. Discuss
this claim.

---

To evaluate the portfolio manager's claim, we need to assess performance
relative to risk, which is typically done using the Capital Asset Pricing Model
(CAPM).

### Expected return under CAPM

The CAPM equation:

$$
\text{Expected return} = R_f + \beta (R_m - R_f)
$$

Where:

* $R_f = 5\%$
* $\beta = 0.2$
* $R_m = -30\%$

The expected return is: $5 + 0.2 \times ( -30\% - 5\%) = -2\%$.

### Actual return vs. expected return

The actual return is -10\%. The expected return is -2\%. This means that the
manager underperformed the risk-adjusted expectation by
$-10\% - (-2\%) = -8\%$.

### Interpretation

Even though the portfolio performed better than the market (-10\% vs. -30\%),
the portfolio took on very little market risk ($\beta = 0.2$). Given that low
risk, it was expected to lose less---only 2\%. But it lost much more than that.

This suggests that the active management strategy failed to deliver on a
risk-adjusted basis. The claim of "good performance" does not hold up under CAPM
analysis and, in fact, the manager's alpha (value added) is negative:

$$
\alpha = \text{Actual return} - \text{expected return} = -8\%
$$

### Conclusion

The manager did better than the market in absolute terms, but much worse than
expected in risk-adjusted terms. Since the whole point of active management is
to outperform after adjusting for risk, the performance was poor, not good.

---

## 25. Futures contracts and a portfolio beta

It is July 16. A company has a portfolio of stocks worth \$100 million. The beta
of the portfolio is 1.2. The company would like to use the December futures
contract on a stock index to change the beta of the portfolio to 0.5 during the
period July 16 to November 16. The index futures price is currently 2,000 and
each contract is on \$250 times the index.

(a) What position should the company take?
(b) Suppose that the company changes its mind and decides to increase the beta
of the portfolio from 1.2 to 1.5. What position in futures contracts should it
take?

---

To solve this, we'll use the formula for adjusting the beta of a portfolio using
index futures

$$
\text{Number of contracts} = \frac{(\beta_T - \beta_P) \times V_P}{V_F}
$$

Where:

* $\beta_T$: Target beta
* $\beta_P$: Current beta
* $V_P$: value of the portfolio
* $V_F$: Value of one futures contract = futures price $\times$ contract
  multiplier

### Given:

* $V_P$: \$100 million
* Futures price: 2,000
* Contract size: \$250 $\times$ index = $\$250 \times 2,000 = \$500,000$
* $V_F$: \$500,000

### A. Reduce the beta from 1.2 to 0.5

$$
\begin{aligned}
\text{Number of contracts} &= \frac{0.5 - 1.2) \times 100,000,000}{500,000}\\
  &= -140
\end{aligned}
$$

The company should short 140 contracts.

### B. Increase the beta from 1.2 to 1.5


$$
\begin{aligned}
\text{Number of contracts} &= \frac{1.5 - 1.2) \times 100,000,000}{500,000}\\
  &= 60
\end{aligned}
$$

The company should go long 60 contracts.

---

# Appendix - Capital asset pricing model

The capital asset pricing model (CAPM) is a model that can be used to relate the
expected return from an asset to the risk of the return. The risk in the return
from an asset is divided into two parts. _Systematic risk_ is the risk related
to the return from the market as a whole and cannot be diversified away.
_Nonsystematic risk_ is risk that is unique to the asset and can be diversified
away by choosing a large portfolio of different assets. CAPM argues that the
return should depend only on systematic risk.

$$
\text{Epected return on asset} = R_F + \beta (R_M - R_F)
$$
